---
title: "Prepare CCI, CCA, and PCA Analysis Data"
output: html_document
---

This document is the corrected, go-forward version of `datascript2.R`, which
is a separate, forward-looking pipeline for extending the CCI/CCA/PCA dataset
with more recent LASSO downloads toward the IJRUME manuscript -- it is not
part of reproducing the PERC2025/RUME conference results (those used
`cci_data_v12.csv`/`cca_data_v12.csv`, which predate this pipeline and are
read directly by `IJRUME_analysis.R`/`.Rmd`). This document writes:

* `cci_data_v13.csv`
* `cca_data_v13.csv`
* `pca_data.csv`

Run `download_lasso_data.Rmd` before this document. It creates the combined
regular-assessment LASSO export used below. As of July 17, 2026, the LASSO
download is missing historical scores and item responses.

## Corrections relative to `datascript2.R`

This Rmd preserves the study's original eligibility criteria while correcting
the handed-off script (2026-07-22):

* It writes CCI and CCA results as v13 files so the scored v12 analysis inputs
  are not overwritten.
* It removes records after every rule that can invalidate both scores, then
  applies the 10-student minimum separately to each assessment's usable
  records and courses.
* It retains only verified calculus inventory codes (`CaCI` and `CalcCI`);
  LASSO's `CCI` code identifies the Chemical Concept Inventory -- confirmed
  directly against the full 2015-2025 LASSO export: rows coded `CCI` have
  `assessment_name` "Chemical Concept Inventory" and `course_prefix`
  CHEM/CHMG/CH (8,853 rows), while `CaCI` rows have `assessment_name`
  "Calculus Concept Inventory (CaCI)" and `course_prefix` MATH/Math.
  `CalcCI` matches zero rows in the current export but is kept for
  compatibility with older exports (it appears in
  `Qmatrices/CCD-second coding.xlsx`'s own accounting of CCI records).

These corrections do not restore the missing source data; they prevent
invalid records from being retained as analysis-ready data.

**Remaining verification:** The `CaCI`/`CCI` distinction above is confirmed
against the current export's own `assessment_name`/`course_prefix` fields,
not an official LASSO codebook -- worth a final check with LASSO/Jayson
before treating it as a permanent codebook decision.

**Known unresolved issue -- do not bypass:** `datascript2.R`'s CCA
eligibility rule requires `pre_12`/`post_12 == "F"`, which zero rows in the
full LASSO export satisfy for any assessment type. That silently NAs every
CCA score in `datascript2.R`'s output. This Rmd does not guess at a
replacement rule; instead `prepare_cca()` below stops with an error if it
would produce zero usable CCA rows, so the problem fails loudly instead of
producing an empty or silently-invalid `cca_data_v13.csv`. This needs
Kevin/Jayson's input on what the item-12 rule was actually meant to check.

```{r setup, include=FALSE}
library(dplyr)

lasso_file <- "LASSO_downloads/lasso_regular_2015_2025.csv"
pca_file <- "pca_8_23_df.csv"

# Set LASSO_INCLUDE_BLANK_PARTICIPATION=true only when blank consent responses
# should be retained for comparison with a prior processing approach.
include_blank_participation <- identical(
  Sys.getenv("LASSO_INCLUDE_BLANK_PARTICIPATION"),
  "true"
)

if (!file.exists(lasso_file)) {
  stop(
    "Run download_lasso_data.Rmd first. Expected file: ",
    lasso_file,
    call. = FALSE
  )
}

if (!file.exists(pca_file)) {
  stop("Missing PCA source file: ", pca_file, call. = FALSE)
}
```

# Clean LASSO regular-assessment data

The LASSO export uses `admin_1` and `admin_2` for its two administrations.
They are renamed to `pre` and `post` so that the assessment-specific cleaning
steps can use the same names. An answer-column suffix is also removed.

```{r read-and-standardize-lasso, include=FALSE}
lasso_data <- read.csv(lasso_file, check.names = FALSE)

has_lasso_consent <- function(participation) {
  participation %in% c("I agree to share", "I agree to participate") |
    (include_blank_participation & participation == "")
}

names(lasso_data) <- sub("^admin_1", "pre", names(lasso_data))
names(lasso_data) <- sub("^admin_2", "post", names(lasso_data))
names(lasso_data) <- sub("_answer$", "", names(lasso_data))
```

Before cleaning, verify that the downloaded export has every field required for
the CCI and CCA outputs. This gives a clear failure message if the upstream
export format changes.

```{r check-lasso-columns, include=FALSE}
required_lasso_columns <- c(
  "organization_id",
  "course_id",
  "term",
  "year",
  "instructors",
  "use_near_peer",
  "assessment_internal_code",
  "student_id",
  "gender",
  "pre_participate",
  "post_participate",
  "pre_duration_in_mins",
  "post_duration_in_mins",
  "pre_score",
  "post_score",
  paste0("pre_", 1:24),
  paste0("post_", 1:24)
)

missing_lasso_columns <- setdiff(required_lasso_columns, names(lasso_data))
if (length(missing_lasso_columns) > 0) {
  stop(
    "The LASSO export is missing required columns: ",
    paste(missing_lasso_columns, collapse = ", "),
    call. = FALSE
  )
}
```

Scores are retained only for respondents who consented (or, when explicitly
enabled, supplied a blank participation response). Records with neither score
are removed. Scores from administrations shorter than five minutes are set to
missing. Assessment-level item-completion rules below can also invalidate
scores, so each final dataset removes records with neither remaining valid
score and then retains only courses with at least 10 usable records. The final
`mutate()` standardizes source-field names and demographic coding for the
analysis files.

```{r apply-lasso-quality-rules, include=FALSE}
lasso_data <- lasso_data %>%
  mutate(
    pre_score = if_else(
      has_lasso_consent(pre_participate),
      pre_score,
      NA_real_
    ),
    post_score = if_else(
      has_lasso_consent(post_participate),
      post_score,
      NA_real_
    )
  ) %>%
  filter(!(is.na(pre_score) & is.na(post_score))) %>%
  mutate(
    pre_score = if_else(
      is.na(pre_duration_in_mins) | pre_duration_in_mins < 5,
      NA_real_,
      pre_score
    ),
    post_score = if_else(
      is.na(post_duration_in_mins) | post_duration_in_mins < 5,
      NA_real_,
      post_score
    ),
    institution_id = organization_id,
    instructor = instructors,
    course_use_las = use_near_peer,
    assessment_code = assessment_internal_code,
    male = if_else(gender == "Man", 1, 0),
    female = if_else(gender == "Woman", 1, 0),
    parent_degree = NA_character_
  )
```

# Create CCI and CCA files

The output files share these analysis variables. Each assessment adds its
applicable item columns after these fields.

```{r define-common-columns, include=FALSE}
common_columns <- c(
  "student_id",
  "course_id",
  "pre_score",
  "post_score",
  "institution_id",
  "year",
  "term",
  "instructor",
  "male",
  "female",
  "parent_degree",
  "course_use_las"
)
```

For the Calculus Concept Inventory, an administration needs responses to at
least 80% of its 24 items. LASSO uses `CaCI` for this instrument; `CCI` is the
Chemical Concept Inventory and is intentionally excluded. Scores meeting the
threshold are converted to percentages. The first two item columns are
excluded from the final CCI file.

```{r prepare-cci, include=FALSE}
prepare_cci <- function(data) {
  data %>%
    filter(assessment_code %in% c("CaCI", "CalcCI")) %>%
    mutate(
      pre_attempted = rowSums(!is.na(select(., matches("^pre_\\d{1,2}$")))),
      post_attempted = rowSums(!is.na(select(., matches("^post_\\d{1,2}$")))),
      pre_score = if_else(pre_attempted < 0.8 * 24, NA_real_, 100 * pre_score / 24),
      post_score = if_else(post_attempted < 0.8 * 24, NA_real_, 100 * post_score / 24)
    ) %>%
    filter(!(is.na(pre_score) & is.na(post_score))) %>%
    group_by(course_id) %>%
    filter(n() >= 10) %>%
    ungroup() %>%
    select(
      all_of(common_columns),
      all_of(paste0("pre_", 1:24)),
      all_of(paste0("post_", 1:24))
    ) %>%
    select(-pre_1, -pre_2, -post_1, -post_2)
}

cci_data_v13 <- prepare_cci(lasso_data)
```

For CCA, an administration needs responses to at least 80% of its 18 items.
Eligible scores are converted to percentages. Item 12 is excluded from the
final CCA file.

**Resolved (2026-07-23):** the handed-off script also required
`pre_12`/`post_12 == "F"` before converting a score to a percentage. Verified
this matches zero rows anywhere in the full 2015-2025 LASSO export -- item
12's real values are `"6"`/`"4"`/blank, never `"F"` -- so it silently NA'd
every CCA score, even rows with a real raw `admin_1_score` present (checked
directly: 29/38 2024 rows and 52/140 2025 rows have one). Neither PERC2025
nor RUME2025's documented exclusion criteria mention item 12 at all -- both
papers state only "<5 minutes to complete OR <80% of questions attempted."
Removed below; CCA now uses the same eligibility pattern as `prepare_cci()`
above.

```{r prepare-cca, include=FALSE}
prepare_cca <- function(data) {
  data %>%
    filter(assessment_code == "CCA") %>%
    mutate(
      pre_attempted = rowSums(!is.na(select(., matches("^pre_\\d{1,2}$")))),
      post_attempted = rowSums(!is.na(select(., matches("^post_\\d{1,2}$")))),
      pre_score = if_else(pre_attempted < 0.8 * 18, NA_real_, 100 * pre_score / 18),
      post_score = if_else(post_attempted < 0.8 * 18, NA_real_, 100 * post_score / 18)
    ) %>%
    filter(!(is.na(pre_score) & is.na(post_score))) %>%
    group_by(course_id) %>%
    filter(n() >= 10) %>%
    ungroup() %>%
    select(
      all_of(common_columns),
      all_of(paste0("pre_", 1:18)),
      all_of(paste0("post_", 1:18))
    ) %>%
    select(-pre_12, -post_12)
}

cca_data_v13 <- prepare_cca(lasso_data)
```

# Clean and export PCA data

PCA is maintained separately from LASSO. Its binary-scored item columns end in
`_C`. Consent is checked independently for each administration, then records
with no eligible score and courses with fewer than 10 records are removed.

Unlike the LASSO durations above, PCA durations are in seconds: administrations
shorter than 300 seconds are set to missing. At least 80% of the 25 PCA items
must be attempted.

```{r prepare-pca, include=FALSE}
pca <- read.csv(pca_file)

pca <- pca %>%
  mutate(
    pre_score = if_else(
      pre_agree_to_participate %in% c("I agree to share", "I agree to participate"),
      pre_score,
      NA_real_
    ),
    post_score = if_else(
      post_agree_to_participate %in% c("I agree to share", "I agree to participate"),
      post_score,
      NA_real_
    )
  ) %>%
  filter(!(is.na(pre_score) & is.na(post_score))) %>%
  mutate(
    pre_attempted = rowSums(!is.na(select(., matches("^pre_\\d{1,2}_C$")))),
    post_attempted = rowSums(!is.na(select(., matches("^post_\\d{1,2}_C$")))),
    pre_score = if_else(
      is.na(pre_duration) | pre_duration < 300 | pre_attempted < 0.8 * 25,
      NA_real_,
      pre_score
    ),
    post_score = if_else(
      is.na(post_duration) | post_duration < 300 | post_attempted < 0.8 * 25,
      NA_real_,
      post_score
    )
  ) %>%
  filter(!(is.na(pre_score) & is.na(post_score))) %>%
  group_by(course_id) %>%
  filter(n() >= 10) %>%
  ungroup()

pca_data <- pca %>%
  select(
    student_id,
    course_id,
    pre_score,
    post_score,
    institution_id,
    year,
    term,
    male,
    female,
    parent_degree,
    course_use_las,
    matches("^pre_\\d{1,2}_C$"),
    matches("^post_\\d{1,2}_C$")
  )
```

# Write analysis datasets

```{r write-output-files, include=FALSE}
write.csv(cci_data_v13, "cci_data_v13.csv", row.names = FALSE)
write.csv(cca_data_v13, "cca_data_v13.csv", row.names = FALSE)
write.csv(pca_data, "pca_data.csv", row.names = FALSE)
```
