---
title: "Download LASSO Data for the CCD Study"
output: html_document
---

# Purpose

This document downloads the LASSO **regular** assessment exports used to
prepare the Calculus Concept Inventory (CCI) and Calculus Concept Assessment
(CCA) datasets for this study. It makes the same export request as the supplied Postman collection, but does
**not** require the Postman desktop application or its JSON files.

The files downloaded here are raw LASSO exports. They are not yet the
analysis-ready `cci_data_v12.csv` and `cca_data_v12.csv` files used by
`IJRUME_analysis.R`. Run `datascript2.R` after downloading to apply the
study's consent, completion-time, item-attempt, and assessment-specific
cleaning rules.

# Before you begin

## Software and R packages

Use R 4.1 or newer and RStudio, if available. Install the required packages
once on each computer:

```{r install-packages, eval=FALSE}
install.packages("httr2")
```

## No Postman files required

This Rmd contains the LASSO production API and UI addresses required for the
export request. It can be shared and run as a single file:

```text
your-project/
└── download_lasso_data.Rmd
```

The API and UI addresses are not credentials. This document contains no API
key, password, token, or student-level data. If LASSO later requires
authentication, do **not** add credentials to a shared Rmd; store them in a
local configuration file or use the access method supplied by LASSO.

The download directory contains student-level data. Store the project in an
approved location, do not share the downloaded CSV files outside approved data
procedures, and do not commit them to a public repository.

## Run the document

1. Open `download_lasso_data.Rmd` in RStudio.
2. Edit the `years` and `terms` settings in the **Choose exports** section if
   needed.
3. Click **Knit** to generate the HTML report, or run all code chunks
   interactively.
4. Find the annual CSV files and combined file in `LASSO_downloads/`.

The first download may take several minutes because each year is requested
separately.

# LASSO request configuration

```{r setup, message=FALSE}
required_packages <- "httr2"
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the required packages first: install.packages(c(",
    paste(sprintf("'%s'", missing_packages), collapse = ", "),
    "))",
    call. = FALSE
  )
}

library(httr2)

# Production settings obtained from the LASSO Postman environment and a
# successful export request. These are public service addresses.
api <- "https://tools-api.lassoeducation.org"
origin <- "https://tools.lassoeducation.org"
content_type <- "text/csv"
```

# Define the download request

The function below requests one year of regular LASSO assessments. The term
parameter `1,2,3,4` requests all four academic terms.

```{r download-functions}
download_lasso_regular_export <- function(year, terms = 1:4, destination) {
  request(paste0(api, "/api/assessments/", year, "/regular/export")) |>
    req_url_query(term = paste(terms, collapse = ",")) |>
    req_headers(
      Origin = origin,
      `Content-Type` = content_type
    ) |>
    req_perform() |>
    resp_body_raw() |>
    writeBin(destination)

  destination
}
```

# Choose exports

The current study's LASSO data span 2015 through 2025. Change `years` to a
smaller range, such as `2025`, to download a single year. To request selected
terms only, change `terms` to values such as `c(1, 3)`.

```{r download-exports}
years <- 2015:2025
terms <- 1:4
download_directory <- "LASSO_downloads"
dir.create(download_directory, showWarnings = FALSE, recursive = TRUE)

export_files <- vapply(
  years,
  function(year) {
    destination <- file.path(
      download_directory,
      sprintf("lasso_regular_%s.csv", year)
    )
    download_lasso_regular_export(
      year,
      terms = terms,
      destination = destination
    )
  },
  character(1)
)

export_files
```

# Combine annual exports

The annual LASSO exports can have different columns across years. This section
keeps every column found in any annual file, adds an `export_year` column for
traceability, and writes one combined CSV. Values unavailable in a particular
year are recorded as `NA`.

```{r combine-exports}
# Combine annual exports while retaining the export year for traceability.
# Align the union of columns because LASSO export schemas can differ by year.
annual_exports <- lapply(export_files, function(file) {
    data <- read.csv(file, check.names = FALSE)
    data$export_year <- as.integer(
      sub(".*lasso_regular_([0-9]{4})\\.csv$", "\\1", file)
    )
    data
  })

all_columns <- unique(unlist(lapply(annual_exports, names)))
annual_exports <- lapply(annual_exports, function(data) {
  missing_columns <- setdiff(all_columns, names(data))
  data[missing_columns] <- NA
  data[all_columns]
})

lasso_regular_data <- do.call(rbind, annual_exports)

combined_file <- file.path(
  download_directory,
  sprintf("lasso_regular_%s_%s.csv", min(years), max(years))
)
write.csv(lasso_regular_data, combined_file, row.names = FALSE)

data.frame(
  files = c(export_files, combined_file),
  rows = c(
    vapply(annual_exports, nrow, integer(1)),
    nrow(lasso_regular_data)
  )
)
```

# Output and next steps

The document creates:

| File | Description |
|---|---|
| `LASSO_downloads/lasso_regular_YYYY.csv` | Raw regular-assessment export for one year |
| `LASSO_downloads/lasso_regular_START_END.csv` | Combined raw export, including `export_year` |

These files include more than the CCI and CCA records and are not directly
used by `IJRUME_analysis.R`. To reproduce the study analysis:

1. Update `datascript2.R` so its raw LASSO input path points to the applicable
   file in `LASSO_downloads/`. The script currently points to legacy data
   locations and filenames.
2. Ensure the separate PCA source file, `pca_8_23_df.csv`, is also available;
   it is not downloaded from LASSO by this document.
3. Run `datascript2.R` to create `cci_data_v12.csv`, `cca_data_v12.csv`, and
   `pca_data.csv`.
4. Run `IJRUME_analysis.R`.

# Troubleshooting

| Problem | Resolution |
|---|---|
| `Install the required packages first` | Run the installation command in **Software and R packages**, then restart R. |
| HTTP 500 or another HTTP error | Confirm that the production addresses in **LASSO request configuration** are current. If the problem persists, contact the LASSO administrator with the year, assessment type (`regular`), and error status. |
| Certificate or network error | Connect through the network normally used for LASSO access and contact the LASSO administrator or local IT support. Do not disable certificate validation. |
| Analysis script cannot find cleaned data | Run `datascript2.R` after downloading; the raw LASSO export is not the final analysis format. |
