---
title: "CCI, CCA, and PCA Analysis"
date: "2026-07-17"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  run_models: false
  run_cdm: false
  qmatrix_path: "Qmatrix.xlsx"
  write_outputs: false
  export_sankey_images: false
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

required_packages <- c(
  "dplyr", "tidyr", "readxl", "ggplot2", "mirt", "GDINA",
  "networkD3", "patchwork", "htmlwidgets", "webshot2", "magick",
  "lmerTest", "stringr", "tibble"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}
invisible(lapply(required_packages, library, character.only = TRUE))

# `params` is supplied by rmarkdown::render(); this fallback also permits
# an in-memory knitr syntax check.
if (!exists("params")) {
  params <- list(
    run_models = FALSE,
    run_cdm = FALSE,
    qmatrix_path = "Qmatrix.xlsx",
    write_outputs = FALSE,
    export_sankey_images = FALSE
  )
}

set.seed(1234)
run_models <- isTRUE(params$run_models)
run_cdm <- isTRUE(params$run_cdm)
qmatrix_path <- params$qmatrix_path
write_outputs <- isTRUE(params$write_outputs)
export_sankey_images <- isTRUE(params$export_sankey_images)
```

# Purpose and inputs

This report reproduces the meaningful descriptive, multilevel, item-response,
and cognitive-diagnostic workflows from `IJRUME_analysis.R` in a documented
order. `IJRUME_analysis.R` is the legacy script behind the analyses reported
in:

* Roberge et al., "Calculus Cognitive Diagnostic: Mathematics skills tested
  in first semester calculus courses" (PERC 2025;
  `Articles/PERC2025_Roberge.pdf`)
* Roberge et al., "Moving Beyond Total Scores with Skill-Proficiency Profiles
  on Concept Inventories" (RUME conference proceedings;
  `Articles/RUME 2025 CCD-UNMASKED-PRESUBMISSION.pdf`)

This Rmd is the updated, go-forward version being built toward the IJRUME
manuscript. `IJRUME_analysis.R` itself has also been patched (2026-07-22) to
the minimum needed to run and reproduce those two manuscripts' fit
statistics -- see the header comment in that file for the exact list of
fixes and its one remaining known bug (the Sankey/profile section can still
crash or silently misalign for respondents GDINA excludes). Everything below
is the corrected version of that same analysis.

## Corrections relative to `IJRUME_analysis.R`

This Rmd corrects the original script's broken CCA/CCI/PCA cognitive-diagnostic
objects, inconsistent Q-matrix paths, undefined PCA missingness variables, and
mislabeling of CCA MAP estimates as EAP. It uses the same item-scoring rules
for each assessment, validates Q-matrix dimensions and values before fitting,
and retains respondent-row alignment when GDINA excludes insufficient item
responses. It also corrects the longitudinal 3PL parameter-table handling,
labels the course effect size as `d_av`, and reports repeated student-course
pairs and student/course crossings rather than silently treating them as
independent observations.

The following CSV files must be in the working directory:

* `cci_data_v12.csv` — CCI student records, scores, and items `pre_3` through
  `pre_24` and `post_3` through `post_24`;
* `cca_data_v12.csv` — CCA student records, scores, and items `pre_1` through
  `pre_18` (without item 12) and their post-test counterparts;
* `pca_data.csv` — PCA student records, scores, and 25 binary `_C` items for
  each test occasion.

**Version requirement:** This analysis deliberately reads
`cci_data_v12.csv` and `cca_data_v12.csv`.  Do **not** silently substitute
the v13 files produced by `datascript2.R`/`datascript2.Rmd`: the newly
downloaded raw exports lack historic scores needed by this analysis.  Version
12 is retained specifically to preserve those historic CCI and CCA scores.

The cognitive-diagnostic/Sankey portion additionally requires
the workbook named by `qmatrix_path`, with sheets named `pca`, `cca`, and
`cci`. The default workbook, `Qmatrix.xlsx`, is
**not present in this project**.  `Qmatrix.csv` is present, but it is a
single 20-row, 8-attribute matrix and cannot be substituted safely for the
three worksheet-specific matrices (the analyses require 25, 17, and 22 item
rows, respectively). Each worksheet must be ordered exactly like the item
columns described above; GDINA matches Q-matrix rows to item columns by
position. Thus, the Sankey section is reported as unavailable unless the
workbook is supplied.

For a quick, non-destructive render, leave all parameters at their defaults.
Set `run_models: true` to estimate the HLM and IRT models. Set
`run_cdm: true` to run only the cognitive-diagnostic/Sankey section with the
selected workbook. Set
`write_outputs: true` only when the course and item-parameter CSV outputs
should be written. Sankey image export is separately opt-in because it writes
HTML and PNG files.

The input summary reports repeated `student_id`/`course_id` pairs but does not
remove them automatically. They may represent legitimate repeated
administrations; their treatment requires a documented study decision before
model fitting.

The Rmd uses GDINA consistently for all three instruments, unlike the original
script's mixed DINA/GDINA choices. Confirm that this is the intended model
family before treating resulting profiles as confirmatory. The real
Q-matrix's item order and content also remain required for substantive results;
the synthetic example workbook is for software practice only.

```{r load-data}
cci_data <- read.csv("cci_data_v12.csv", stringsAsFactors = FALSE)
cca_data <- read.csv("cca_data_v12.csv", stringsAsFactors = FALSE)
pca_data <- read.csv("pca_data.csv", stringsAsFactors = FALSE)

summarize_repeated_student_course_pairs <- function(data, assessment) {
  repeated_pairs <- data |>
    select(student_id, course_id) |>
    count(student_id, course_id, name = "records") |>
    filter(records > 1)
  tibble::tibble(
    assessment = assessment,
    input_rows = nrow(data),
    repeated_student_course_pairs = nrow(repeated_pairs),
    rows_in_repeated_pairs = sum(repeated_pairs$records)
  )
}

input_sizes <- tibble::tibble(
  assessment = c("CCI", "CCA", "PCA"),
  students = c(nrow(cci_data), nrow(cca_data), nrow(pca_data)),
  variables = c(ncol(cci_data), ncol(cca_data), ncol(pca_data))
)
input_sizes
repeated_pair_summary <- bind_rows(
  summarize_repeated_student_course_pairs(cci_data, "CCI"),
  summarize_repeated_student_course_pairs(cca_data, "CCA"),
  summarize_repeated_student_course_pairs(pca_data, "PCA")
)
repeated_pair_summary
```

# Descriptive statistics

## Reusable summaries

The helpers below create consistent score, gain, course-level, and
demographic summaries without duplicating the same transformations for each
assessment.

```{r descriptive-helpers}
make_score_data <- function(data, assessment) {
  bind_rows(
    transmute(data, score = pre_score, version = "Pre"),
    transmute(data, score = post_score, version = "Post")
  ) |>
    mutate(assessment = assessment)
}

summarize_scores <- function(score_data) {
  score_data |>
    group_by(assessment, version) |>
    summarise(
      mean_score = mean(score, na.rm = TRUE),
      median_score = median(score, na.rm = TRUE),
      sd_score = sd(score, na.rm = TRUE),
      n = sum(!is.na(score)),
      .groups = "drop"
    )
}

summarize_courses <- function(data) {
  data |>
    group_by(course_id) |>
    summarise(
      mean_pre = mean(pre_score, na.rm = TRUE),
      mean_post = mean(post_score, na.rm = TRUE),
      sd_pre = sd(pre_score, na.rm = TRUE),
      sd_post = sd(post_score, na.rm = TRUE),
      n_pre = sum(!is.na(pre_score)),
      n_post = sum(!is.na(post_score)),
      .groups = "drop"
    ) |>
    mutate(
      gain = round(mean_post - mean_pre, 2),
      normgain = if_else(
        mean_pre == 100, NA_real_,
        round((mean_post - mean_pre) / (100 - mean_pre), 3)
      ),
      average_sd = (sd_pre + sd_post) / 2,
      d_av = if_else(
        average_sd == 0, NA_real_,
        round((mean_post - mean_pre) / average_sd, 3)
      )
    ) |>
    select(-average_sd)
}

summarize_demographics <- function(data, include_instructor = FALSE) {
  result <- list(
    courses = count(data, course_id, name = "n"),
    sex = count(data, male, female, name = "n")
  )
  if (include_instructor) {
    result$instructors <- count(data, instructor, name = "n")
  }
  result
}

write_csv_if_requested <- function(data, path) {
  if (write_outputs) {
    write.csv(data, path, row.names = FALSE)
  }
  invisible(data)
}
```

```{r descriptive-data}
score_data <- bind_rows(
  make_score_data(cci_data, "CCI"),
  make_score_data(cca_data, "CCA"),
  make_score_data(pca_data, "PCA")
)

gain_data <- bind_rows(
  transmute(cci_data, assessment = "CCI", gain = post_score - pre_score),
  transmute(cca_data, assessment = "CCA", gain = post_score - pre_score),
  transmute(pca_data, assessment = "PCA", gain = post_score - pre_score)
)

score_summary <- summarize_scores(score_data)
score_summary

course_summaries <- list(
  cci = summarize_courses(cci_data),
  cca = summarize_courses(cca_data),
  pca = summarize_courses(pca_data)
)

demographic_summaries <- list(
  cci = summarize_demographics(cci_data, include_instructor = TRUE),
  cca = summarize_demographics(cca_data, include_instructor = TRUE),
  pca = summarize_demographics(pca_data)
)

write_csv_if_requested(demographic_summaries$cci$courses, "cci_courses.csv")
write_csv_if_requested(demographic_summaries$cci$instructors, "cci_inst.csv")
write_csv_if_requested(demographic_summaries$cca$courses, "cca_courses.csv")
write_csv_if_requested(demographic_summaries$cca$instructors, "cca_inst.csv")
write_csv_if_requested(demographic_summaries$pca$courses, "pca_courses.csv")
write_csv_if_requested(course_summaries$cci, "course_means_cci.csv")
write_csv_if_requested(course_summaries$cca, "course_means_cca.csv")
write_csv_if_requested(course_summaries$pca, "course_means_pca.csv")
```

## Score and gain distributions

```{r descriptive-plots}
score_density_plots <- lapply(c("CCI", "CCA", "PCA"), function(assessment_name) {
  ggplot(
    filter(score_data, .data$assessment == assessment_name),
    aes(x = score, fill = version)
  ) +
    geom_density(na.rm = TRUE, alpha = 0.3, color = "black") +
    labs(title = paste(assessment_name, "pre- and post-test scores")) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
})
wrap_plots(score_density_plots, ncol = 3)

ggplot(gain_data, aes(x = gain, fill = assessment)) +
  geom_density(na.rm = TRUE, alpha = 0.3, color = "black") +
  geom_vline(xintercept = 0, linetype = "dotted", linewidth = 1) +
  labs(title = "Pre/post gains by assessment", x = "Post score − pre score") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
```

## Course-level variation

Course-level summaries are retained as data outputs and as distributions of
the standardized pre/post difference. `d_av` uses the pre/post average
standard deviation and does not incorporate within-student correlation; it is
not the paired-samples `d_z`. A missing or zero pooled standard deviation
yields `NA` rather than an infinite effect size.

```{r course-level-plots}
course_summaries$cci
course_summaries$cca
course_summaries$pca

par(mfrow = c(1, 3))
for (assessment in c("cci", "cca", "pca")) {
  hist(
    course_summaries[[assessment]]$d_av,
    main = paste(toupper(assessment), "course d_av"),
    xlab = "d_av"
  )
}
par(mfrow = c(1, 1))
```

# Multilevel pre/post models

Scores are reshaped to one row per student/test occasion.  The null model
partitions variance among students, courses, and residual error; the
fixed-occasion model estimates the average post-test change; and the final
model allows the pre/post change to vary across courses. This treats student
and course effects as crossed, so a repeated `student_id` is assumed to
identify the same student across every associated course.

```{r hlm-helpers}
make_long_scores <- function(data) {
  data |>
    select(student_id, course_id, pre_score, post_score) |>
    pivot_longer(
      cols = c(pre_score, post_score),
      names_to = "version",
      values_to = "score"
    ) |>
    mutate(version = if_else(version == "pre_score", 0, 1)) |>
    filter(!is.na(score))
}

calculate_iccs <- function(model) {
  variance_components <- as.data.frame(VarCorr(model))
  component_variance <- function(group) {
    variance_components$vcov[variance_components$grp == group]
  }
  total_variance <- sum(
    component_variance("student_id"),
    component_variance("course_id"),
    component_variance("Residual")
  )
  tibble::tibble(
    student_icc = component_variance("student_id") / total_variance,
    course_icc = component_variance("course_id") / total_variance
  )
}

summarize_student_course_crossing <- function(data, assessment) {
  data |>
    distinct(student_id, course_id) |>
    count(student_id, name = "courses_per_student") |>
    summarise(
      assessment = assessment,
      students = n(),
      students_in_multiple_courses = sum(courses_per_student > 1),
      .groups = "drop"
    )
}

fit_hlm_models <- function(data) {
  long_data <- make_long_scores(data)
  null_model <- lmerTest::lmer(
    score ~ (1 | student_id) + (1 | course_id),
    data = long_data,
    REML = TRUE
  )
  fixed_change_model <- lmerTest::lmer(
    score ~ version + (1 | student_id) + (1 | course_id),
    data = long_data,
    REML = FALSE
  )
  random_change_model <- lmerTest::lmer(
    score ~ version + (1 | student_id) + (1 + version | course_id),
    data = long_data,
    REML = FALSE
  )
  list(
    data = long_data,
    null_model = null_model,
    fixed_change_model = fixed_change_model,
    random_change_model = random_change_model,
    iccs = calculate_iccs(null_model),
    comparison = anova(fixed_change_model, random_change_model)
  )
}
```

```{r student-course-diagnostics}
bind_rows(
  summarize_student_course_crossing(cci_data, "CCI"),
  summarize_student_course_crossing(cca_data, "CCA"),
  summarize_student_course_crossing(pca_data, "PCA")
)
```

```{r hlm-models, eval=run_models}
hlm_results <- list(
  CCI = fit_hlm_models(cci_data),
  CCA = fit_hlm_models(cca_data),
  PCA = fit_hlm_models(pca_data)
)

lapply(hlm_results, `[[`, "iccs")
lapply(hlm_results, `[[`, "comparison")
```

# Item-response theory

## Item scoring and matrices

CCI and CCA responses are lowercased and scored from the answer keys.  A
student with no score for an occasion has all item responses for that
occasion set to missing.  PCA already provides binary `_C` item columns.
This preserves partially observed pre/post records for IRT estimation while
removing records missing both occasions.

```{r irt-helpers}
cci_pre_key <- c(
  pre_3 = "e", pre_4 = "e", pre_5 = "a", pre_6 = "e", pre_7 = "c",
  pre_8 = "c", pre_9 = "b", pre_10 = "e", pre_11 = "b", pre_12 = "a",
  pre_13 = "b", pre_14 = "a", pre_15 = "a", pre_16 = "d", pre_17 = "c",
  pre_18 = "a", pre_19 = "b", pre_20 = "a", pre_21 = "d", pre_22 = "c",
  pre_23 = "a", pre_24 = "a"
)
cci_post_key <- setNames(unname(cci_pre_key), sub("^pre_", "post_", names(cci_pre_key)))

cca_pre_key <- c(
  pre_1 = "b", pre_2 = "b", pre_3 = "b", pre_4 = "c", pre_5 = "a",
  pre_6 = "c", pre_7 = "d", pre_8 = "c", pre_9 = "b", pre_10 = "d",
  pre_11 = "b", pre_13 = "d", pre_14 = "b", pre_15 = "b", pre_16 = "d",
  pre_17 = "a", pre_18 = "c"
)
cca_post_key <- setNames(unname(cca_pre_key), sub("^pre_", "post_", names(cca_pre_key)))

score_items <- function(data, pre_key, post_key) {
  response_columns <- c(names(pre_key), names(post_key))
  data |>
    mutate(across(all_of(response_columns), tolower)) |>
    mutate(
      across(
        all_of(names(pre_key)),
        ~ as.integer(.x == pre_key[cur_column()])
      ),
      across(
        all_of(names(post_key)),
        ~ as.integer(.x == post_key[cur_column()])
      )
    )
}

mask_unobserved_occasions <- function(data, pre_items, post_items) {
  data |>
    mutate(
      across(all_of(pre_items), ~ replace(.x, is.na(pre_score), NA)),
      across(all_of(post_items), ~ replace(.x, is.na(post_score), NA))
    ) |>
    filter(!(is.na(pre_score) & is.na(post_score)))
}

prepare_irt_data <- function(data, pre_items, post_items) {
  response_data <- data |>
    select(student_id, all_of(pre_items), all_of(post_items))
  response_matrix <- response_data |>
    select(-student_id)
  response_matrix[] <- lapply(response_matrix, function(x) {
    x <- as.numeric(x)
    ifelse(is.na(x), NA_integer_, as.integer(x))
  })
  list(
    ids = response_data$student_id,
    responses = response_matrix,
    pre_items = pre_items,
    post_items = post_items
  )
}

make_longitudinal_model <- function(number_of_items) {
  mirt::mirt.model(paste0(
    "F1 = 1-", number_of_items, "\n",
    "F2 = ", number_of_items + 1, "-", 2 * number_of_items, "\n",
    "COV = F1*F2"
  ))
}

cci_pre_items <- names(cci_pre_key)
cci_post_items <- names(cci_post_key)
cca_pre_items <- names(cca_pre_key)
cca_post_items <- names(cca_post_key)
pca_pre_items <- paste0("pre_", 1:25, "_C")
pca_post_items <- paste0("post_", 1:25, "_C")

cci_scored <- score_items(cci_data, cci_pre_key, cci_post_key)
cca_scored <- score_items(cca_data, cca_pre_key, cca_post_key)

cci_clean <- mask_unobserved_occasions(cci_scored, cci_pre_items, cci_post_items)
cca_clean <- mask_unobserved_occasions(cca_scored, cca_pre_items, cca_post_items)
pca_clean <- mask_unobserved_occasions(pca_data, pca_pre_items, pca_post_items)

irt_inputs <- list(
  CCI = prepare_irt_data(cci_clean, cci_pre_items, cci_post_items),
  CCA = prepare_irt_data(cca_clean, cca_pre_items, cca_post_items),
  PCA = prepare_irt_data(pca_clean, pca_pre_items, pca_post_items)
)
```

## Invariant longitudinal 3PL models

For each assessment, corresponding pre/post items share discrimination,
difficulty, and guessing parameters. A logit-normal prior with mean `-1.23`
and standard deviation `0.50` approximates the intended center and spread of
the former `Beta(5, 17)` guessing prior while using mirt's recommended
parameterization for a bounded guessing parameter. CCI and PCA use EAP
ability scores; the original CCA workflow used MAP scores, which is retained
here and labeled accordingly.

```{r model-helpers}
find_parameter <- function(values, item, parameter) {
  which(values$item == item & values$name == parameter)
}

make_invariance_constraints <- function(values, pre_items, post_items) {
  unlist(
    Map(
      function(pre_item, post_item) {
        list(
          c(find_parameter(values, pre_item, "a1"),
            find_parameter(values, post_item, "a2")),
          c(find_parameter(values, pre_item, "d"),
            find_parameter(values, post_item, "d")),
          c(find_parameter(values, pre_item, "g"),
            find_parameter(values, post_item, "g"))
        )
      },
      pre_items, post_items
    ),
    recursive = FALSE
  )
}

apply_guessing_prior <- function(values, item_names) {
  for (item in item_names) {
    guessing_row <- find_parameter(values, item, "g")
    values$prior.type[guessing_row] <- "norm"
    values$prior_1[guessing_row] <- -1.23
    values$prior_2[guessing_row] <- 0.50
  }
  values
}

fit_invariant_3pl <- function(irt_input, score_method = "EAP") {
  number_of_items <- length(irt_input$pre_items)
  model <- make_longitudinal_model(number_of_items)
  values <- mirt::mirt(
    data = irt_input$responses,
    model = model,
    itemtype = "3PL",
    pars = "values"
  )
  values <- apply_guessing_prior(
    values,
    c(irt_input$pre_items, irt_input$post_items)
  )
  fit <- mirt::mirt(
    data = irt_input$responses,
    model = model,
    itemtype = "3PL",
    pars = values,
    constrain = make_invariance_constraints(
      values, irt_input$pre_items, irt_input$post_items
    ),
    SE = TRUE,
    method = "MHRM",
    technical = list(NCYCLES = 3000)
  )

  ability <- mirt::fscores(fit, method = score_method)
  theta <- tibble::tibble(
    student_id = irt_input$ids,
    theta_pre = ability[, "F1"],
    theta_post = ability[, "F2"],
    d_theta = theta_post - theta_pre
  )
  covariance <- coef(fit, simplify = TRUE)$cov
  model_correlation <- covariance[1, 2] / sqrt(covariance[1, 1] * covariance[2, 2])
  score_correlation <- cor(
    theta$theta_pre, theta$theta_post, use = "pairwise.complete.obs"
  )

  item_parameters <- coef(fit, IRTpars = TRUE, simplify = TRUE)$items |>
    as.data.frame() |>
    mutate(
      a1 = round(as.numeric(a1), 3),
      b = round(as.numeric(b), 3),
      g = round(as.numeric(g), 3)
    )

  pre_observed <- rowSums(!is.na(irt_input$responses[, irt_input$pre_items, drop = FALSE])) > 0
  post_observed <- rowSums(!is.na(irt_input$responses[, irt_input$post_items, drop = FALSE])) > 0

  list(
    fit = fit,
    theta = theta,
    complete_case_theta = theta[pre_observed & post_observed, , drop = FALSE],
    model_correlation = model_correlation,
    score_correlation = score_correlation,
    score_method = score_method,
    fit_statistic = mirt::M2(fit),
    item_parameters = item_parameters
  )
}

plot_irt_results <- function(results, assessment) {
  complete_case_correlation <- cor(
    results$complete_case_theta$theta_pre,
    results$complete_case_theta$theta_post,
    use = "pairwise.complete.obs"
  )
  ability_plot <- ggplot(results$theta, aes(theta_pre, theta_post)) +
    geom_point(alpha = 0.4, size = 1) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
    labs(
      title = paste(assessment, "(invariant 3PL): pre vs. post ability"),
      subtitle = paste0(
        "Model r = ", round(results$model_correlation, 3),
        " | ", results$score_method, " r = ",
        round(results$score_correlation, 3)
      ),
      x = expression(theta[pre]),
      y = expression(theta[post])
    ) +
    theme_minimal()
  change_plot <- ggplot(results$theta, aes(d_theta)) +
    geom_histogram(bins = 30) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
      title = paste("Change in ability (post − pre):", assessment),
      subtitle = paste0(
        results$score_method, " r for students observed at both occasions = ",
        round(complete_case_correlation, 3)
      ),
      x = expression(Delta * theta),
      y = "Count"
    ) +
    theme_minimal()
  list(ability = ability_plot, change = change_plot)
}

plot_item_parameters <- function(item_parameters, assessment) {
  par(mfrow = c(1, 3))
  for (parameter in c("a1", "b", "g")) {
    hist(
      item_parameters[[parameter]],
      breaks = "FD",
      col = "gray",
      main = paste(assessment, parameter),
      xlab = parameter
    )
    abline(v = median(item_parameters[[parameter]], na.rm = TRUE), lty = 2)
  }
  par(mfrow = c(1, 1))
}
```

```{r irt-models, eval=run_models}
irt_results <- list(
  CCI = fit_invariant_3pl(irt_inputs$CCI, score_method = "EAP"),
  CCA = fit_invariant_3pl(irt_inputs$CCA, score_method = "MAP"),
  PCA = fit_invariant_3pl(irt_inputs$PCA, score_method = "EAP")
)

for (assessment in names(irt_results)) {
  results <- irt_results[[assessment]]
  print(results$fit_statistic)
  print(plot_irt_results(results, assessment)$ability)
  print(plot_irt_results(results, assessment)$change)
  plot_item_parameters(results$item_parameters, assessment)
  write_csv_if_requested(
    results$item_parameters,
    paste0("cf_", tolower(assessment), ".csv")
  )
}
```

# Cognitive-diagnostic profiles and Sankey diagrams

The original exploratory Sankey code was consolidated into the functions
below and corrected to use the scored CCI/CCA response matrices.  This
section is skipped until the required workbook exists; it is also skipped
when `run_cdm` is false. GDINA requires at least two valid item responses
for an occasion, so respondents below that threshold are retained in the
pre/post cohort but labeled `Missing` for that occasion before transitions are
counted.

```{r sankey-helpers}
has_qmatrix <- file.exists(qmatrix_path)

prepare_cd_items <- function(data, item_columns, occasion) {
  data |>
    select(all_of(item_columns)) |>
    setNames(sub(paste0("^", occasion, "_([0-9]+).*"), "q\\1", item_columns))
}

extract_attribute_estimates <- function(fit) {
  as.matrix(GDINA::personparm(fit, what = "EAP"))
}

validate_qmatrix <- function(qmatrix, item_count, attribute_count, sheet) {
  if (nrow(qmatrix) != item_count || ncol(qmatrix) != attribute_count) {
    stop(
      "Q-matrix sheet '", sheet, "' must have ", item_count, " rows and ",
      attribute_count, " attribute columns after metadata columns are removed."
    )
  }
  if (!all(unlist(qmatrix, use.names = FALSE) %in% c(0, 1))) {
    stop("Q-matrix sheet '", sheet, "' must contain only binary 0/1 entries.")
  }
  qmatrix
}

make_profile_labels <- function(fit, retained_rows, total_rows) {
  attribute_estimates <- extract_attribute_estimates(fit)
  if (nrow(attribute_estimates) != length(retained_rows)) {
    stop(
      "GDINA returned a different number of profiles than the supplied ",
      "respondent rows; profiles cannot be aligned safely."
    )
  }

  binary_attributes <- ifelse(attribute_estimates >= 0.5, 1L, 0L)
  profile_labels <- apply(binary_attributes, 1, paste0, collapse = "")
  profile_labels[apply(is.na(attribute_estimates), 1, any)] <- "Uncertain"

  # GDINA excludes rows with fewer than two valid responses. Restore them to
  # the original student-row order as missing profiles before pairing pre/post.
  labels <- rep("Missing", total_rows)
  labels[retained_rows] <- profile_labels
  labels
}

make_sankey <- function(pre_labels, post_labels, minimum_n = 10) {
  transitions <- tibble(
    pre_class = paste("Pre:", pre_labels),
    post_class = paste("Post:", post_labels)
  ) |>
    add_count(pre_class, name = "n_pre") |>
    add_count(post_class, name = "n_post") |>
    filter(n_pre >= minimum_n, n_post >= minimum_n) |>
    count(pre_class, post_class, name = "value")

  nodes <- data.frame(
    name = sort(unique(c(transitions$pre_class, transitions$post_class)))
  )
  links <- transitions |>
    transmute(
      source = match(pre_class, nodes$name) - 1,
      target = match(post_class, nodes$name) - 1,
      value
    )
  networkD3::sankeyNetwork(
    Links = links,
    Nodes = nodes,
    Source = "source",
    Target = "target",
    Value = "value",
    NodeID = "name",
    fontSize = 16,
    nodeWidth = 30,
    width = 800
  )
}

load_qmatrix <- function(sheet, item_count, attribute_count, columns_to_drop = "LO") {
  readxl::read_excel(qmatrix_path, sheet = sheet) |>
    select(-any_of(columns_to_drop)) |>
    validate_qmatrix(item_count, attribute_count, sheet)
}

fit_profile_sankey <- function(clean_data, pre_items, post_items, qmatrix_sheet,
                               attribute_count,
                               qmatrix_columns_to_drop = "LO") {
  qmatrix <- load_qmatrix(qmatrix_sheet, length(pre_items), attribute_count, qmatrix_columns_to_drop)
  pre_data <- prepare_cd_items(clean_data, pre_items, "pre")
  post_data <- prepare_cd_items(clean_data, post_items, "post")

  # Match GDINA's documented minimum of two valid item responses per person.
  pre_rows <- which(rowSums(!is.na(pre_data)) >= 2)
  post_rows <- which(rowSums(!is.na(post_data)) >= 2)
  if (!length(pre_rows) || !length(post_rows)) {
    stop("No respondents have enough valid item responses for GDINA.")
  }

  pre_fit <- GDINA::GDINA(
    dat = pre_data[pre_rows, , drop = FALSE], Q = qmatrix, model = "GDINA"
  )
  post_fit <- GDINA::GDINA(
    dat = post_data[post_rows, , drop = FALSE], Q = qmatrix, model = "GDINA"
  )
  pre_labels <- make_profile_labels(pre_fit, pre_rows, nrow(clean_data))
  post_labels <- make_profile_labels(post_fit, post_rows, nrow(clean_data))
  list(
    plot        = make_sankey(pre_labels, post_labels),
    qmatrix     = qmatrix,
    pre_fit     = pre_fit,
    post_fit    = post_fit,
    pre_rows    = pre_rows,
    post_rows   = post_rows,
    pre_labels  = pre_labels,
    post_labels = post_labels
  )
}
```

```{r sankey-models, eval=run_cdm}
if (!has_qmatrix) {
  cat(
    "`Qmatrix.xlsx` is absent, so the cognitive-diagnostic and Sankey analyses ",
    "were not run.  See the input requirements above.\n",
    sep = ""
  )
} else {
  sankey_results <- list(
    PCA = fit_profile_sankey(
      pca_clean, pca_pre_items, pca_post_items, "pca",
      attribute_count = 4,
      qmatrix_columns_to_drop = c("LO", "D")
    ),
    CCA = fit_profile_sankey(
      cca_clean, cca_pre_items, cca_post_items, "cca",
      attribute_count = 5
    ),
    CCI = fit_profile_sankey(
      cci_clean, cci_pre_items, cci_post_items, "cci",
      attribute_count = 5
    )
  )
  lapply(sankey_results, `[[`, "plot")
}
```

## Optional Sankey image export

This export mirrors the original PCA/CCI side-by-side image workflow.  It is
not run by default because it writes `pca_sankey.html`, `cci_sankey.html`,
their PNG screenshots, and `sankeys_side_by_side.png`.

```{r export-sankey-images, eval=run_cdm && export_sankey_images}
stopifnot(exists("sankey_results"))

htmlwidgets::saveWidget(sankey_results$PCA$plot, "pca_sankey.html", selfcontained = TRUE)
htmlwidgets::saveWidget(sankey_results$CCI$plot, "cci_sankey.html", selfcontained = TRUE)
webshot2::webshot("pca_sankey.html", file = "pca_sankey.png",
                  vwidth = 900, vheight = 600, zoom = 2)
webshot2::webshot("cci_sankey.html", file = "cci_sankey.png",
                  vwidth = 900, vheight = 600, zoom = 2)

pca_image <- magick::image_read("pca_sankey.png") |> magick::image_trim()
cci_image <- magick::image_read("cci_sankey.png") |> magick::image_trim()
target_height <- max(
  magick::image_info(pca_image)$height,
  magick::image_info(cci_image)$height
)
pca_image <- magick::image_extent(
  pca_image,
  geometry = paste0(magick::image_info(pca_image)$width, "x", target_height),
  gravity = "center"
)
cci_image <- magick::image_extent(
  cci_image,
  geometry = paste0(magick::image_info(cci_image)$width, "x", target_height),
  gravity = "center"
)
spacer <- magick::image_blank(width = 100, height = target_height, color = "white")
magick::image_append(c(pca_image, spacer, cci_image), stack = FALSE) |>
  magick::image_write("sankeys_side_by_side.png")
```

# Current tables for the IJRUME manuscript

These are the numbers this project will actually use going forward,
computed via the corrected pipeline above (all-GDINA, invariant 3PL,
row-alignment-fixed Sankey) on the full current data -- no year-filtering,
no published-value comparison. For a comparison against PERC2025/RUME2025's
published numbers using year-filtered data and the original DINA/GDINA
methodology, see `PERC_RUME_Reproduction.Rmd`.

## Descriptive statistics

```{r current-descriptives}
score_summary |> filter(version == "Post")

summarize_matched_pairs <- function(data, assessment) {
  matched <- data |> filter(!is.na(pre_score), !is.na(post_score))
  tibble::tibble(
    assessment = assessment,
    n = nrow(matched),
    mean_pre = round(mean(matched$pre_score), 1),
    sd_pre = round(sd(matched$pre_score), 1),
    mean_post = round(mean(matched$post_score), 1),
    sd_post = round(sd(matched$post_score), 1)
  ) |>
    mutate(d_av = round((mean_post - mean_pre) / ((sd_pre + sd_post) / 2), 2))
}

bind_rows(
  summarize_matched_pairs(cci_data, "CCI"),
  summarize_matched_pairs(cca_data, "CCA"),
  summarize_matched_pairs(pca_data, "PCA")
)
```

## Item distribution by skill

```{r current-item-distribution, eval=has_qmatrix}
skill_order <- c("P", "L", "D", "AD", "I")

item_counts <- function(q, assessment) {
  counts <- colSums(q)
  full <- setNames(rep(0L, length(skill_order)), skill_order)
  full[names(counts)] <- as.integer(counts)
  tibble::tibble(assessment = assessment, skill = skill_order, n_items = full)
}

bind_rows(
  item_counts(load_qmatrix("cca", length(cca_pre_items), 5), "CCA"),
  item_counts(load_qmatrix("cci", length(cci_pre_items), 5), "CCI"),
  item_counts(load_qmatrix("pca", length(pca_pre_items), 4, c("LO", "D")), "PCA")
) |>
  pivot_wider(names_from = assessment, values_from = n_items) |>
  mutate(Total = CCA + CCI + PCA)
```

## Fit statistics and classification accuracy

```{r current-fit-stats, eval=run_cdm && has_qmatrix}
fit_stats <- function(fit) {
  mf <- GDINA::modelfit(fit)
  tibble::tibble(RMSEA = round(mf$RMSEA2, 3), SRMSR = round(mf$SRMSR, 3))
}

bind_rows(
  mutate(fit_stats(sankey_results$CCA$post_fit), assessment = "CCA"),
  mutate(fit_stats(sankey_results$CCI$post_fit), assessment = "CCI"),
  mutate(fit_stats(sankey_results$PCA$post_fit), assessment = "PCA")
) |> select(assessment, RMSEA, SRMSR)

# GDINA::CA()$tau_k labels attributes generically ("A1".."An" in Q-matrix
# column order); relabel using the Q-matrix each fit was built from.
ca_row <- function(result) {
  ca <- GDINA::CA(result$post_fit)$tau_k
  names(ca) <- names(result$qmatrix)
  sapply(skill_order, function(s) unname(round(ca[s], 2)))
}

tibble::tibble(
  skill = skill_order,
  CCA = ca_row(sankey_results$CCA),
  CCI = ca_row(sankey_results$CCI),
  PCA = ca_row(sankey_results$PCA)
)
```

## HLM model summary

```{r current-hlm-table, eval=run_models}
extract_hlm_table <- function(hlm_result, assessment) {
  coefs <- summary(hlm_result$fixed_change_model)$coefficients
  tibble::tibble(
    assessment = assessment,
    term = c("Intercept", "Post-test"),
    coefficient = round(coefs[, "Estimate"], 2),
    se = round(coefs[, "Std. Error"], 2),
    df = round(coefs[, "df"], 2),
    t = round(coefs[, "t value"], 2),
    p = coefs[, "Pr(>|t|)"]
  )
}

bind_rows(
  extract_hlm_table(hlm_results$CCA, "CCA"),
  extract_hlm_table(hlm_results$CCI, "CCI"),
  extract_hlm_table(hlm_results$PCA, "PCA")
)
```

## Sankey diagrams

```{r current-sankey-display, eval=run_cdm && has_qmatrix}
sankey_results$PCA$plot
sankey_results$CCI$plot
sankey_results$CCA$plot
```
