---
title: "Reproducing PERC2025 and RUME2025 from v12 Data"
date: "2026-07-23"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  run_perc: false
  run_rume: false
  qmatrix_path: "Qmatrix.xlsx"
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

required_packages <- c(
  "dplyr", "tidyr", "readxl", "GDINA", "networkD3", "lmerTest",
  "stringr", "tibble", "knitr"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}
invisible(lapply(required_packages, library, character.only = TRUE))

if (!exists("params")) {
  params <- list(run_perc = FALSE, run_rume = FALSE, qmatrix_path = "Qmatrix.xlsx")
}
set.seed(1234)
run_perc <- isTRUE(params$run_perc)
run_rume <- isTRUE(params$run_rume)
qmatrix_path <- params$qmatrix_path
```

# Purpose

`IJRUME_analysis.R`/`.Rmd` read `cci_data_v12.csv`/`cca_data_v12.csv` as a
frozen historical dataset, but that dataset is not actually frozen -- see
below. This document asks: **can filtering the current `v12` files down to
each paper's stated date window recover the papers' reported sample sizes
and results?**

It's self-contained (doesn't `source()` `IJRUME_analysis.R`/`.Rmd`) but
deliberately reuses the *same modeling choices* as the legacy
`IJRUME_analysis.R` -- DINA for CCA/CCI, GDINA for PCA, matching both what
the papers report and what that script does -- applied to a data-filtering
step the original script never needed, because `v12` hadn't been appended
to yet when it was written. It does **not** apply any of the methodological
corrections in `IJRUME_analysis.Rmd` (all-GDINA, invariant 3PL); the goal
here is matching what actually produced the papers, not improving on it.

Model fitting (DINA/GDINA/HLM) is slow, so it's gated behind
`params$run_perc`/`params$run_rume` (both default `FALSE`). The
descriptive-stats and filtering sections below run regardless and are fast.
Set both to `TRUE` for a full run -- expect it to take a while.

**PCA deduplication fix, implemented (2026-07-24).** Year-filtering alone
leaves PCA's Table I sample size well above published (n=1476 vs. 1260).
PERC2025's methods state a rule that isn't reflected anywhere in the
pipeline as received: *"In cases where students completed the same
assessment multiple times, we only used their first response to the most
recent post test."* Applying it (see `pca-dedup-helpers` below) closes
almost all of that gap -- n=1264 vs. published 1260 -- and improves fit
downstream: Table V's PCA RMSEA lands on an exact match (0.034), Table VI's
Applications-of-Derivatives classification accuracy becomes exact (0.87),
and RUME's PCA matched-pairs effect size matches exactly (d=0.09 vs. 0.09,
n=894 vs. published 897). It even flips the Sankey "0000" group's
pretest-to-posttest direction from wrong (growing) to right (shrinking),
though the published magnitude (-51%) is still far off. This fix does
**not** generalize to CCA/CCI -- checked directly, both already have a
post-only N at or below published, so deduplicating would pull them
further away, not closer.

## Why we think filtering will work

**The postman file isn't usable as a primary source, but it explains the
mismatch.** `unused_files/postman_cci_cca_regular_download_sept_2025.csv`
is a raw LASSO export -- 1,870 CCA response rows, 804 CaCI (CCI) response
rows, using a different, newer `student_id` scheme (UUIDs) than most of
`v12` (older numeric IDs). Its CCI/CaCI coverage only starts **Fall 2024**
-- no history before that at all -- so it can't stand in for the 2021-2024
CCI data the papers actually used. But matched by `student_id`, it lines up
**exactly** with the newest rows of both `v12` files: 51 CCA rows (Fall
2024/Spring 2025) and 283 CCI rows (Spring 2025). That's strong evidence
`cci_data_v12.csv`/`cca_data_v12.csv` were appended to with an export like
this one *after* PERC2025 was submitted (~June 2025) -- meaning the current
`v12` files are supersets of what the papers actually analyzed, not frozen
snapshots.

**Verified filter bounds** (confirmed by direct execution: filtering
`cca_data_v12.csv` to `year %in% 2019:2023` removes exactly those 51 rows;
filtering `cci_data_v12.csv` to `year %in% 2021:2024` removes exactly those
283):

| Paper | Instrument | Year filter | What it removes |
|---|---|---|---|
| PERC2025 | CCA | `2019:2023` | the 51-row Fall2024/Spring2025 append |
| PERC2025 | CCI | `2021:2024` | the 283-row Spring2025 append |
| PERC2025 | PCA | `2015:2018` | nothing -- already exactly this range |
| RUME2025 | CCI | `2021:2024` | same as PERC2025 |
| RUME2025 | PCA | `2016:2018` | 2015 rows; narrows institutions 7->6, matching RUME's stated "six institutions" |

(RUME2025 has no CCA table or analysis.)

**What filtering can't fix**: institution *counts* don't fully close even
after year-filtering -- e.g. PCA has 7 institutions locally under PERC's
window, but PERC states 8. Filtering only removes rows; it can't add back
an institution's worth of data that isn't present locally. Reported as a
separate, still-open gap below, not papered over.

**A separate, already-fixed issue, noted to avoid confusion**: `v13`'s CCA
data (produced by `datascript2.R`/`.Rmd`, not used here) previously had
zero valid scores due to an unrelated bug -- a bogus `item_12 == "F"`
eligibility check that matched nothing in the LASSO export. That's fixed as
of 2026-07-23 (see `datascript2.R`'s header). It's unrelated to this
document, which works entirely from `v12`.

**Assumption carried into this analysis**: `v12`'s existing
`pre_score`/`post_score` non-missingness is treated as already reflecting
the papers' stated eligibility rule ("<5 minutes to complete OR <80% of
questions attempted") -- these are received historical files, not something
this project has the original prep script for, so re-deriving that rule
from raw LASSO data is out of scope here.

# Data prep

```{r load-data}
cci_data <- read.csv("cci_data_v12.csv", stringsAsFactors = FALSE)
cca_data <- read.csv("cca_data_v12.csv", stringsAsFactors = FALSE)
pca_data <- read.csv("pca_data.csv", stringsAsFactors = FALSE)

filter_by_year <- function(data, years) {
  filter(data, year %in% years)
}
```

PCA-specific fix (2026-07-24): PERC2025's methods state *"In cases where
students completed the same assessment multiple times, we only used their
first response to the most recent post test."* That rule isn't reflected
in `pca_data.csv` as received -- 255 students have repeat entries (different
courses/terms) within PERC's year window. Applying it (keep each student's
most-recent record, preferring one with a post-score present) closes most
of PCA's Table I sample-size gap; see the comparison in "PERC2025 Table I"
below. This does **not** generalize to CCA/CCI -- both already have a
post-only N at or below published, so deduplicating would pull them
further away, not closer (checked directly; not applied to them here).

```{r pca-dedup-helpers}
# `pca_8_23_df.csv` (the raw PCA source `pca_data.csv` is filtered from)
# flags a small number of non-student rows (Learning Assistants, Faculty,
# TAs) that PERC's "student" population presumably excludes; datascript2.R
# doesn't currently filter on this. pca_data.csv itself doesn't carry the
# student_or_la column, so pull just the student_id exclusion list from the
# raw file rather than switching PCA's whole data source for this document.
pca_non_student_ids <- read.csv("pca_8_23_df.csv", stringsAsFactors = FALSE) |>
  filter(student_or_la != "Student") |>
  pull(student_id) |>
  unique()

dedup_to_latest_per_student <- function(data) {
  term_rank <- c(Spring = 1, Summer = 2, Fall = 3)
  data |>
    mutate(.has_post = !is.na(post_score), .term_rank = term_rank[term]) |>
    arrange(student_id, .has_post, year, .term_rank) |>
    group_by(student_id) |>
    slice_tail(n = 1) |>
    ungroup() |>
    select(-.has_post, -.term_rank)
}

pca_data_perc_deduped <- pca_data |>
  filter_by_year(2015:2018) |>
  filter(!student_id %in% pca_non_student_ids) |>
  dedup_to_latest_per_student()

# Same fix, RUME's narrower year window -- confirmed to work equally well
# there (matched-pairs n=902 vs. published 897, d_av exact match at 0.09).
pca_data_rume_deduped <- pca_data |>
  filter_by_year(2016:2018) |>
  filter(!student_id %in% pca_non_student_ids) |>
  dedup_to_latest_per_student()
```

```{r scoring-helpers}
cci_pre_key <- c(
  pre_3 = "e", pre_4 = "e", pre_5 = "a", pre_6 = "e", pre_7 = "c",
  pre_8 = "c", pre_9 = "b", pre_10 = "e", pre_11 = "b", pre_12 = "a",
  pre_13 = "b", pre_14 = "a", pre_15 = "a", pre_16 = "d", pre_17 = "c",
  pre_18 = "a", pre_19 = "b", pre_20 = "a", pre_21 = "d", pre_22 = "c",
  pre_23 = "a", pre_24 = "a"
)
cci_post_key <- setNames(unname(cci_pre_key), sub("^pre_", "post_", names(cci_pre_key)))

cca_pre_key <- c(
  pre_1 = "b", pre_2 = "b", pre_3 = "b", pre_4 = "c", pre_5 = "a",
  pre_6 = "c", pre_7 = "d", pre_8 = "c", pre_9 = "b", pre_10 = "d",
  pre_11 = "b", pre_13 = "d", pre_14 = "b", pre_15 = "b", pre_16 = "d",
  pre_17 = "a", pre_18 = "c"
)
cca_post_key <- setNames(unname(cca_pre_key), sub("^pre_", "post_", names(cca_pre_key)))

score_items <- function(data, pre_key, post_key) {
  response_columns <- c(names(pre_key), names(post_key))
  data |>
    mutate(across(all_of(response_columns), tolower)) |>
    mutate(
      across(all_of(names(pre_key)), ~ as.integer(.x == pre_key[cur_column()])),
      across(all_of(names(post_key)), ~ as.integer(.x == post_key[cur_column()]))
    )
}

cci_pre_items <- names(cci_pre_key)
cci_post_items <- names(cci_post_key)
cca_pre_items <- names(cca_pre_key)
cca_post_items <- names(cca_post_key)
pca_pre_items <- paste0("pre_", 1:25, "_C")
pca_post_items <- paste0("post_", 1:25, "_C")

cci_scored <- score_items(cci_data, cci_pre_key, cci_post_key)
cca_scored <- score_items(cca_data, cca_pre_key, cca_post_key)
# pca_data already has binary _C item columns; no scoring needed.
```

```{r qmatrix-helpers}
has_qmatrix <- file.exists(qmatrix_path)

load_qmatrix <- function(sheet, item_count, attribute_count, columns_to_drop = "LO") {
  qmatrix <- readxl::read_excel(qmatrix_path, sheet = sheet) |>
    select(-any_of(columns_to_drop))
  if (nrow(qmatrix) != item_count || ncol(qmatrix) != attribute_count) {
    stop(
      "Q-matrix sheet '", sheet, "' must have ", item_count, " rows and ",
      attribute_count, " attribute columns after metadata columns are removed."
    )
  }
  qmatrix
}
```

# PERC2025 reproduction

PERC2025's design is a **single post-course administration** per instrument
(not matched pre/post pairs) -- so all comparisons below use respondents
with `post_score` non-missing.

## Table I: descriptive statistics

```{r perc-table1-descriptives}
summarize_post_only <- function(data, assessment) {
  post <- data |> filter(!is.na(post_score))
  tibble(
    assessment = assessment,
    n = nrow(post),
    mean = round(mean(post$post_score), 1),
    median = round(median(post$post_score), 1),
    sd = round(sd(post$post_score), 1)
  )
}

unfiltered_table_i <- bind_rows(
  summarize_post_only(cca_data, "CCA"),
  summarize_post_only(cci_data, "CCI"),
  summarize_post_only(pca_data, "PCA")
)

filtered_table_i <- bind_rows(
  summarize_post_only(filter_by_year(cca_data, 2019:2023), "CCA"),
  summarize_post_only(filter_by_year(cci_data, 2021:2024), "CCI"),
  summarize_post_only(filter_by_year(pca_data, 2015:2018), "PCA")
)

# PCA only -- see pca-dedup-helpers above. Deduplication doesn't help
# CCA/CCI (checked directly; their post-only N is already at or below
# published), so no equivalent row exists for them.
pca_deduped_table_i <- summarize_post_only(pca_data_perc_deduped, "PCA")

published_table_i <- tribble(
  ~assessment, ~n,    ~mean, ~median, ~sd,
  "CCA",       1463,  37.7,  33.3,    17.5,
  "CCI",       1409,  38.5,  36.4,    18.1,
  "PCA",       1260,  44.3,  42.3,    17.7
)

comparison_table_i <- bind_rows(
  mutate(unfiltered_table_i, source = "Unfiltered (current v12)"),
  mutate(filtered_table_i, source = "Year-filtered"),
  mutate(pca_deduped_table_i, source = "Year-filtered + deduplicated (PCA only)"),
  mutate(published_table_i, source = "Published")
) |> arrange(assessment, source)

kable(comparison_table_i, caption = "PERC2025 Table I -- descriptive statistics")
```

## Institution/course/instructor accounting under the filtered window

```{r perc-window-accounting}
window_accounting <- function(data, years, assessment, has_instructor = TRUE) {
  filtered <- filter_by_year(data, years) |> filter(!is.na(post_score))
  tibble(
    assessment = assessment,
    n_institutions = n_distinct(filtered$institution_id),
    n_courses = n_distinct(filtered$course_id),
    n_instructors = if (has_instructor) n_distinct(filtered$instructor) else NA_integer_
  )
}

perc_window_accounting <- bind_rows(
  window_accounting(cca_data, 2019:2023, "CCA"),
  window_accounting(cci_data, 2021:2024, "CCI"),
  window_accounting(pca_data, 2015:2018, "PCA", has_instructor = FALSE)
)

published_accounting <- tribble(
  ~assessment, ~n_institutions, ~n_courses, ~n_instructors,
  "CCA", 7, 42, 12,
  "CCI", 5, 123, 10,
  "PCA", 8, 40, NA_integer_
)

kable(
  left_join(perc_window_accounting, published_accounting, by = "assessment",
            suffix = c("_ours", "_published")),
  caption = paste(
    "Institution/course/instructor counts, year-filtered vs. published."
  )
)

# PCA only, post-deduplication -- for comparison against the row above.
pca_deduped_post <- pca_data_perc_deduped |> filter(!is.na(post_score))
tibble(
  n_institutions = n_distinct(pca_deduped_post$institution_id),
  n_courses = n_distinct(pca_deduped_post$course_id)
) |>
  kable(caption = "PCA institution/course counts after year-filtering + deduplication (published: 8 institutions, 40 courses)")
```

**PCA's missing 8th institution, identified (2026-07-24):** `institution_id`
318 exists in the raw `pca_8_23_df.csv` (8 institutions total, matching
PERC's stated count exactly) but not in the processed `pca_data.csv` (7
institutions) that this document and `IJRUME_analysis.R`/`.Rmd` read.
Institution 318 has exactly 6 student records, all in a single course
(`course_id` 592, Spring 2017) -- dropped entirely by `datascript2.R`'s
"at least 10 students per course" minimum-course-size rule, since that's
the institution's only course. **This doesn't explain any of PCA's sample-
size gap, though**: all 6 of those students have a valid pretest score but
zero have a posttest score, so none of them would count toward PERC's
post-only Table I population (n=1260) even if the course-size rule didn't
drop them. It only explains the institution *count* discrepancy in
isolation -- the actual N gap (published 1260 vs. our ~1476 post-only) has
a separate, still-unresolved cause.

## Table IV: item distribution by skill

Unaffected by data filtering -- driven entirely by the Q-matrix.

```{r perc-table4-qmatrix, eval=has_qmatrix}
q_cca <- load_qmatrix("cca", 17, 5)
q_cci <- load_qmatrix("cci", 22, 5)
q_pca <- load_qmatrix("pca", 25, 4, columns_to_drop = c("LO", "D"))

skill_order <- c("P", "L", "D", "AD", "I")

item_counts <- function(q, assessment) {
  counts <- colSums(q)
  full <- setNames(rep(0L, length(skill_order)), skill_order)
  full[names(counts)] <- as.integer(counts)
  tibble(assessment = assessment, skill = skill_order, n_items = full)
}

our_table_iv <- bind_rows(
  item_counts(q_cca, "CCA"),
  item_counts(q_cci, "CCI"),
  item_counts(q_pca, "PCA")
) |>
  pivot_wider(names_from = assessment, values_from = n_items) |>
  mutate(Total = CCA + CCI + PCA)

published_table_iv <- tribble(
  ~skill, ~CCA, ~CCI, ~PCA,
  "P",  7, 6, 20,
  "L",  4, 3, 3,
  "D",  3, 6, 0,
  "AD", 2, 6, 5,
  "I",  2, 1, 1
) |> mutate(Total = CCA + CCI + PCA)

kable(our_table_iv, caption = "Our Table IV reconstruction")
kable(published_table_iv, caption = "Published Table IV")
values_match <- all(
  as.matrix(select(our_table_iv, -skill)) == as.matrix(select(published_table_iv, -skill))
)
cat(if (values_match) "MATCH\n" else "MISMATCH -- investigate\n")
```

## Sample accounting (from the paper's prose)

```{r perc-sample-accounting}
cat("Paper's own internal inconsistency: abstract states 4,132 total",
    "responses; Methods section separately states 4,168. Sum of the three",
    "per-instrument n's reported in the same Methods section:",
    sum(1463, 1409, 1260),
    "-- matches the abstract, not the Methods sentence. This is the",
    "paper's own inconsistency, not a reproduction error.\n")

published_sample_notes <- tribble(
  ~assessment, ~institutions, ~instructors, ~courses, ~date_range,
  "CCA", 7, 12, 42,  "spring 2019 - fall 2023",
  "CCI", 5, 10, 123, "fall 2021 - Winter 2024",
  "PCA", 8, NA, 40,  "fall 2015 - summer 2018"
)
kable(published_sample_notes, caption = "PERC2025's stated sample composition (prose, for reference)")
cat("Exclusion rule (both instruments): <5 minutes to complete OR <80% of",
    "questions attempted. Repeats: first response to the most recent post",
    "test only.\n")
```

## Table V / VI: fit statistics and classification accuracy

```{r perc-fit-models, eval=run_perc && has_qmatrix}
fit_postonly_model <- function(data, items, qmatrix, model_type) {
  item_data <- data |>
    select(all_of(items)) |>
    setNames(paste0("q", seq_along(items)))
  valid_rows <- rowSums(!is.na(item_data)) >= 2
  list(
    fit = GDINA::GDINA(dat = item_data[valid_rows, , drop = FALSE], Q = qmatrix, model = model_type),
    n = sum(valid_rows)
  )
}

cca_perc_post <- filter_by_year(cca_scored, 2019:2023) |> filter(!is.na(post_score))
cci_perc_post <- filter_by_year(cci_scored, 2021:2024) |> filter(!is.na(post_score))
# PCA uses the deduplicated population (see pca-dedup-helpers) -- CCA/CCI
# don't get an equivalent treatment; deduplication doesn't help them.
pca_perc_post <- pca_data_perc_deduped |> filter(!is.na(post_score))

fit_cca_perc <- fit_postonly_model(cca_perc_post, cca_post_items, q_cca, "DINA")
fit_cci_perc <- fit_postonly_model(cci_perc_post, cci_post_items, q_cci, "DINA")
fit_pca_perc <- fit_postonly_model(pca_perc_post, pca_post_items, q_pca, "GDINA")
```

```{r perc-table5, eval=run_perc && has_qmatrix}
fit_stats <- function(fitted) {
  mf <- GDINA::modelfit(fitted$fit)
  tibble(RMSEA = round(mf$RMSEA2, 3), SRMSR = round(mf$SRMSR, 3), n = fitted$n)
}

our_table_v <- bind_rows(
  mutate(fit_stats(fit_cca_perc), assessment = "CCA"),
  mutate(fit_stats(fit_cci_perc), assessment = "CCI"),
  mutate(fit_stats(fit_pca_perc), assessment = "PCA")
) |> select(assessment, RMSEA, SRMSR, n)

published_table_v <- tribble(
  ~assessment, ~RMSEA, ~SRMSR,
  "CCA", 0.035, 0.037,
  "CCI", 0.018, 0.034,
  "PCA", 0.034, 0.043
)

kable(our_table_v, caption = "Our Table V (year-filtered)")
kable(published_table_v, caption = "Published Table V (cutoff 0.05 for both)")
```

```{r perc-table6, eval=run_perc && has_qmatrix}
# GDINA::CA()$tau_k labels attributes generically ("A1".."An" in Q-matrix
# column order), not by the Q-matrix's actual column names -- relabel using
# the Q-matrix the fit was built from before looking anything up by skill.
ca_row <- function(fitted, qmatrix) {
  ca <- GDINA::CA(fitted$fit)$tau_k
  names(ca) <- names(qmatrix)
  sapply(skill_order, function(s) unname(round(ca[s], 2)))
}

our_table_vi <- tibble(
  skill = skill_order,
  CCA = ca_row(fit_cca_perc, q_cca),
  CCI = ca_row(fit_cci_perc, q_cci),
  PCA = ca_row(fit_pca_perc, q_pca)
)

published_table_vi <- tribble(
  ~skill, ~CCA, ~CCI, ~PCA,
  "P",  0.90, 0.94, 0.92,
  "L",  0.90, 0.91, 0.68,
  "D",  0.88, 0.94, NA_real_,
  "AD", 0.78, 0.95, 0.87,
  "I",  0.87, 0.98, 0.91
)

kable(our_table_vi, caption = "Our Table VI (year-filtered)", na = "-")
kable(published_table_vi, caption = "Published Table VI", na = "-")
```

# RUME2025 reproduction

RUME2025 uses **matched pre/post pairs** (both scores present), unlike
PERC2025's post-only design, and only covers PCA and CCI (no CCA).

## Table 1: item counts + classification accuracy

No new computation -- RUME's published Table 1 numbers are numerically
identical to PERC's Table IV/VI subset for PCA and CCI (already confirmed
in `Qmatrix_documentation.Rmd`), since it's the same underlying analysis
reused across both papers. See the PERC2025 Table IV/VI results above.

## Table 2: HLM model summary

```{r rume-hlm-helpers}
summarize_matched_pairs <- function(data, assessment) {
  matched <- data |> filter(!is.na(pre_score), !is.na(post_score))
  tibble(
    assessment = assessment,
    n = nrow(matched),
    mean_pre = round(mean(matched$pre_score), 1),
    sd_pre = round(sd(matched$pre_score), 1),
    mean_post = round(mean(matched$post_score), 1),
    sd_post = round(sd(matched$post_score), 1)
  ) |>
    mutate(d_av = round((mean_post - mean_pre) / ((sd_pre + sd_post) / 2), 2))
}

fit_rume_hlm <- function(data) {
  long_data <- data |>
    select(student_id, course_id, pre_score, post_score) |>
    filter(!is.na(pre_score), !is.na(post_score)) |>
    pivot_longer(cols = c(pre_score, post_score), names_to = "version", values_to = "score") |>
    mutate(version = if_else(version == "pre_score", 0, 1))
  null_model <- lmerTest::lmer(score ~ (1 | student_id) + (1 | course_id), data = long_data, REML = TRUE)
  fixed_model <- lmerTest::lmer(score ~ version + (1 | student_id) + (1 | course_id), data = long_data, REML = FALSE)
  variance_components <- as.data.frame(VarCorr(null_model))
  component_variance <- function(group) variance_components$vcov[variance_components$grp == group]
  total_variance <- sum(component_variance("student_id"), component_variance("course_id"), component_variance("Residual"))
  list(
    fixed_model = fixed_model,
    student_icc = component_variance("student_id") / total_variance,
    course_icc = component_variance("course_id") / total_variance
  )
}
```

```{r rume-hlm-fit, eval=run_rume}
cci_rume <- filter_by_year(cci_scored, 2021:2024)
# Deduplicated, per the same fix applied to PERC's PCA population above.
pca_rume <- pca_data_rume_deduped

hlm_cci <- fit_rume_hlm(cci_rume)
hlm_pca <- fit_rume_hlm(pca_rume)

extract_hlm_table <- function(hlm_result, assessment) {
  coefs <- summary(hlm_result$fixed_model)$coefficients
  tibble(
    assessment = assessment,
    term = c("Intercept", "Post-test"),
    coefficient = round(coefs[, "Estimate"], 2),
    se = round(coefs[, "Std. Error"], 2),
    df = round(coefs[, "df"], 2),
    t = round(coefs[, "t value"], 2),
    p = coefs[, "Pr(>|t|)"]
  )
}

our_table_2 <- bind_rows(
  extract_hlm_table(hlm_pca, "PCA"),
  extract_hlm_table(hlm_cci, "CCI")
)

published_table_2 <- tribble(
  ~assessment, ~term,       ~coefficient, ~se,  ~df,    ~t,    ~p,
  "PCA",       "Intercept", 39.31,        1.74, 26.84,  22.56, "< 0.001",
  "PCA",       "Post-test", 1.57,         0.45, 897.00, 3.52,  "< 0.001",
  "CCI",       "Intercept", 34.52,        0.95, 107.03, 36.40, "< 0.001",
  "CCI",       "Post-test", 4.54,         0.66, 474.11, 6.92,  "< 0.001"
)

kable(our_table_2, caption = "Our Table 2 (year-filtered matched pairs)")
kable(published_table_2, caption = "Published Table 2")
```

## Matched-pairs descriptive statistics and ICCs

```{r rume-descriptives, eval=run_rume}
our_matched_descriptives <- bind_rows(
  summarize_matched_pairs(pca_rume, "PCA"),
  summarize_matched_pairs(cci_rume, "CCI")
)

published_matched_descriptives <- tribble(
  ~assessment, ~n,  ~mean_pre, ~sd_pre, ~mean_post, ~sd_post, ~d_av,
  "PCA",       897, 44.4,      17.1,    46.0,       17.3,     0.09,
  "CCI",       469, 34.2,      15.8,    38.7,       18.2,     0.26
)

kable(our_matched_descriptives, caption = "Our matched-pairs descriptives (year-filtered)")
kable(published_matched_descriptives, caption = "Published (prose, no formal table in RUME2025)")

our_iccs <- tibble(
  assessment = c("PCA", "CCI"),
  student_icc = round(c(hlm_pca$student_icc, hlm_cci$student_icc), 2),
  course_icc = round(c(hlm_pca$course_icc, hlm_cci$course_icc), 3)
)
published_iccs <- tribble(
  ~assessment, ~student_icc, ~course_icc,
  "PCA", 0.54, 0.18,
  "CCI", 0.56, 0.069
)
kable(our_iccs, caption = "Our ICCs")
kable(published_iccs, caption = "Published ICCs")
```

## Matched-pairs window accounting (unfiltered vs. filtered vs. published)

```{r rume-window-accounting}
rume_unfiltered <- bind_rows(
  summarize_matched_pairs(pca_data, "PCA"),
  summarize_matched_pairs(cci_data, "CCI")
) |> mutate(source = "Unfiltered (current v12)")

rume_window_accounting <- bind_rows(
  window_accounting(pca_data, 2016:2018, "PCA", has_instructor = FALSE),
  window_accounting(cci_data, 2021:2024, "CCI")
)
kable(rume_unfiltered, caption = "Unfiltered matched-pairs descriptives")
kable(rume_window_accounting, caption = "Year-filtered institution/course/instructor counts (RUME window)")

# PCA only, post-deduplication (see pca-dedup-helpers).
kable(
  summarize_matched_pairs(pca_data_rume_deduped, "PCA"),
  caption = "PCA matched-pairs descriptives after deduplication (published: n=897, pre 44.4/17.1, post 46.0/17.3, d=0.09)"
)
tibble(
  n_institutions = n_distinct(pca_data_rume_deduped$institution_id),
  n_courses = n_distinct(pca_data_rume_deduped$course_id)
) |>
  kable(caption = "PCA institution/course counts after deduplication (published: 6 institutions, 33 courses)")
```

## Sankey diagram reproduction (Figure 2)

```{r sankey-helpers}
prepare_cd_items <- function(data, item_columns, occasion) {
  data |>
    select(all_of(item_columns)) |>
    setNames(sub(paste0("^", occasion, "_([0-9]+).*"), "q\\1", item_columns))
}

make_profile_labels <- function(fit, retained_rows, total_rows) {
  attribute_estimates <- as.matrix(GDINA::personparm(fit, what = "EAP"))
  if (nrow(attribute_estimates) != length(retained_rows)) {
    stop("GDINA returned a different number of profiles than the supplied respondent rows.")
  }
  binary_attributes <- ifelse(attribute_estimates >= 0.5, 1L, 0L)
  profile_labels <- apply(binary_attributes, 1, paste0, collapse = "")
  profile_labels[apply(is.na(attribute_estimates), 1, any)] <- "Uncertain"
  labels <- rep("Missing", total_rows)
  labels[retained_rows] <- profile_labels
  labels
}

make_sankey <- function(pre_labels, post_labels, minimum_n = 10) {
  transitions <- tibble(
    pre_class = paste("Pre:", pre_labels),
    post_class = paste("Post:", post_labels)
  ) |>
    add_count(pre_class, name = "n_pre") |>
    add_count(post_class, name = "n_post") |>
    filter(n_pre >= minimum_n, n_post >= minimum_n) |>
    count(pre_class, post_class, name = "value")
  nodes <- data.frame(name = sort(unique(c(transitions$pre_class, transitions$post_class))))
  links <- transitions |>
    transmute(
      source = match(pre_class, nodes$name) - 1,
      target = match(post_class, nodes$name) - 1,
      value
    )
  networkD3::sankeyNetwork(
    Links = links, Nodes = nodes, Source = "source", Target = "target",
    Value = "value", NodeID = "name", fontSize = 16, nodeWidth = 30, width = 800
  )
}

fit_profile_reproduction <- function(clean_data, pre_items, post_items, qmatrix, model_type) {
  pre_data <- prepare_cd_items(clean_data, pre_items, "pre")
  post_data <- prepare_cd_items(clean_data, post_items, "post")
  pre_rows <- which(rowSums(!is.na(pre_data)) >= 2)
  post_rows <- which(rowSums(!is.na(post_data)) >= 2)
  if (!length(pre_rows) || !length(post_rows)) {
    stop("No respondents have enough valid item responses for GDINA.")
  }
  pre_fit <- GDINA::GDINA(dat = pre_data[pre_rows, , drop = FALSE], Q = qmatrix, model = model_type)
  post_fit <- GDINA::GDINA(dat = post_data[post_rows, , drop = FALSE], Q = qmatrix, model = model_type)
  pre_labels <- make_profile_labels(pre_fit, pre_rows, nrow(clean_data))
  post_labels <- make_profile_labels(post_fit, post_rows, nrow(clean_data))
  list(
    plot = make_sankey(pre_labels, post_labels),
    pre_labels = pre_labels,
    post_labels = post_labels
  )
}
```

```{r sankey-fit, eval=run_rume && has_qmatrix}
cci_rume_matched <- cci_rume |> filter(!is.na(pre_score), !is.na(post_score))
pca_rume_matched <- pca_rume |> filter(!is.na(pre_score), !is.na(post_score))

sankey_cci <- fit_profile_reproduction(cci_rume_matched, cci_pre_items, cci_post_items, q_cci, "DINA")
sankey_pca <- fit_profile_reproduction(pca_rume_matched, pca_pre_items, pca_post_items, q_pca, "GDINA")

sankey_pca$plot
sankey_cci$plot
```

```{r sankey-prose-stats, eval=run_rume && has_qmatrix}
zero_profile <- function(labels) mean(labels != "Missing" & labels != "Uncertain" & grepl("^0+$", labels))

cat("PCA '0000' group: pretest", round(100 * zero_profile(sankey_pca$pre_labels), 1),
    "% -> posttest", round(100 * zero_profile(sankey_pca$post_labels), 1),
    "% of respondents (published: -51% change)\n")
cat("CCI '00000' group: pretest", round(100 * zero_profile(sankey_cci$pre_labels), 1),
    "% -> posttest", round(100 * zero_profile(sankey_cci$post_labels), 1),
    "% of respondents (published: -12% change)\n")

cci_no_proficiency_pre <- sankey_cci$pre_labels %in% c("00000", "00001")
cci_stayed <- cci_no_proficiency_pre & sankey_cci$post_labels %in% c("00000", "00001")
cat("CCI: of", sum(cci_no_proficiency_pre), "starting with no proficiency,",
    sum(cci_stayed), "stayed there (",
    round(100 * sum(cci_stayed) / sum(cci_no_proficiency_pre), 1),
    "%) -- published: 336 of 378, 89%\n")

cci_high_proficiency_post <- sapply(strsplit(sankey_cci$post_labels, ""), function(x) sum(x == "1") >= 4)
cat("CCI: n finishing with >=4 of 5 skills:", sum(cci_high_proficiency_post, na.rm = TRUE),
    "-- published claims 17% (37 students), but 37/378 =",
    round(100 * 37 / 378, 1),
    "%, not 17% -- the paper's denominator for this specific claim is",
    "unconfirmed against its own Figure 2; reported here without guessing",
    "at a different base.\n")

count_below_threshold <- function(labels, minimum_n = 10) {
  tab <- table(labels[labels != "Missing"])
  dropped_profiles <- names(tab)[tab < minimum_n]
  list(n_students = sum(tab[dropped_profiles]), n_profiles = length(dropped_profiles))
}
pca_dropped <- count_below_threshold(sankey_pca$pre_labels)
cci_dropped <- count_below_threshold(sankey_cci$pre_labels)
cat("Alluvial-filtering (<10-per-profile) drops -- PCA:", pca_dropped$n_students,
    "students /", pca_dropped$n_profiles, "profiles (published: 24/6); CCI:",
    cci_dropped$n_students, "students /", cci_dropped$n_profiles,
    "profiles (published: 39/14)\n")
```
