---
title: "Q-matrix Documentation, History, and Reconstruction"
date: "2026-07-22"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  write_qmatrix: false
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

required_packages <- c("dplyr", "stringr", "readxl", "writexl", "knitr", "tibble", "tidyr", "purrr")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ", paste(missing_packages, collapse = ", "))
}
invisible(lapply(required_packages, library, character.only = TRUE))

if (!exists("params")) params <- list(write_qmatrix = FALSE)
write_qmatrix <- isTRUE(params$write_qmatrix)
```

# Purpose

This document is the single source of truth for the CCA/CCI/PCA Q-matrix used
by `IJRUME_analysis.R`/`.Rmd`: where it came from, how it was derived from the
project's working files, what it means item-by-item, and what remains
unknown. It **regenerates `Qmatrix.xlsx` from source** (see
[Generating Qmatrix.xlsx]) rather than treating that file as a hand-edited
artifact, so the derivation stays auditable and reproducible if the source
coding workbook changes.

Three instruments are covered:

* **CCA** — Calculus Concept Assessment (17 items)
* **CCI** — Calculus Concept Inventory (22 items)
* **PCA** — Precalculus Concept Assessment (25 items; originally Carlson,
  Oehrtman & Engelke's instrument — see [PCA's original framework: Carlson (2010)])

All three are scored against five broad skills: **P**rerequisites,
**L**imits, **D**erivatives, **A**pplications of **D**erivatives, and
**I**ntegration.

**A note on `Articles/`:** this document cites several PDFs in `Articles/`
by path (the two project manuscripts, plus Carlson (2010) and Epstein
(2013)). That folder is **intentionally not part of this public GitHub
repo** -- it's local reference material only, since it mixes the project's
own manuscripts with copyrighted third-party journal articles and other
material not appropriate to redistribute publicly.

**A note on repo layout:** to keep the repo minimal, only one Q-matrix
source file is tracked: `CCD-second coding.xlsx` (root), which this
document's [Reconstructing the 5-skill Q-matrix] chunk reads directly to
regenerate `Qmatrix.xlsx`. Older/intermediate coding files that are useful
context but not needed to run the analysis (an earlier 39-fine-attribute
coding stage, an old manual LASSO export, an orphaned plotting script) live
in `unused_files/` instead of being deleted outright, since they're
referenced by name in the history below.

**What this document is not:** it does not contain the actual item wording.
See [What's missing: item text] for why, and what to do about it.

# History: how we got to a 5-skill Q-matrix

This chronology is reconstructed from two sources: the `Roberge_agendas/`
meeting-notes documents (Kevin Roberge's and Jayson Nissen's dated working
notes, Jan 2025 -- Mar 2026) and the tab structure of
`CCD-second coding.xlsx` itself, which preserves several coding
iterations side by side.

| When | What happened |
|---|---|
| **Jan 2025** | Kevin starts coding the calculus assessments on LASSO (C1CI, CCAv2/CCA, C2CI, CaCI): "Winnow down LOs to less than 50 skills." Registers for RUME 2025. |
| **Feb 5, 2025** | Confirms LO count needs at least 10 questions per skill to be usable. RUME 2025 paper (submission ID 136, "Moving Beyond Total Scores with Skill-Proficiency Profiles on Concept Inventories") is already under review at this point, already using the 5-skill P/L/D/AD/I scheme per reviewer comments. |
| **Feb 12, 2025** | Builds a much more granular OpenStax-based codebook: ~162 LOs at a fine level, ~36 at a coarser section level; adds P6 (Algebra Basics) and E1 (Interpretation); codes CCI/CaCI/CCA/PCA/C2CI producing 98 item-codes. This is the 27-LO taxonomy reproduced in full below. |
| **Feb 26, 2025** | Literature review pass: Carlson (2010, PCA), Epstein (2013, CCI), Gleason et al. (2015), Carlson et al. (2015). Explores running PCA's DINA model using Carlson's *own* original LOs, not the recoded P/L/D/AD/I scheme. |
| **Mar 7--21, 2025** | Jayson and Kevin add P7 (graphical fluency); build a Q-matrix tab; do an item-by-item KEEP/REMOVE pass on sparsely-coded LOs for CCI and CCA. |
| **Apr 17--25, 2025** | "Codebook is finished" at ~37 fine-grained LOs (P1--P7, L1--L5, D1--D9, AD1--AD10, I1--I7, E1) -- this is what's preserved in `unused_files/Qmatrix_intermediate_coding_stage.xlsx`'s `CCI`/`CCA`/`PCA` tabs. The **`CCD-second coding.xlsx`** workbook is created around this point, with a `2ndCoding Qs` tab holding "actual Q-matrices and suggested revisions" and a Qval-based fit-improvement pass (Qval is a DINA-model diagnostic that flags items whose empirical response pattern doesn't match their assigned Q-matrix row). |
| **May 7, 2025** | Explicit statement of the operative scheme: **"For the CCD we are using five Learning Objectives: P: Prerequisite knowledge, L: Limits, D: Derivatives, AD: Applications of Derivatives, I: Introduction to Integration."** Fit statistics compared across a "full Q-matrix," a "5 LO Q-matrix," and a "5 LO Q-matrix, UPDATED from Qval" (1 change to CCA, 2 changes to CCI) -- this last one is the version reconstructed in this document. |
| **May 21--29, 2025** | "I think what we've done for the PERC paper is fine but I would like to pursue the attribute hierarchies this summer." DINA re-run on a corrected CCA subset; "I have also updated the numbers in the PERC manuscript." |
| **Jun 4, 2025** | PERC2025 paper submitted. |
| **Jun 15--18, 2025** | "CCD on LASSO Update... All IRT, DINA (Guess/Slip) and Q updated." Creates a GitHub repo for the project ("Calc Repo") and shares it with Jayson -- this is the likely origin of the code now in *this* repository. Upload checklist includes `Qmatrix.xlsx` (the 5-LO, Qval-updated version) alongside `Calc_IRT.R`, `calc_data_prep.R`, `Cacl_DINA.R`. |
| **Sep 2025 -- Feb 2026** | Separate RUME conference-proceedings cycle (a second, later RUME submission "recycling" PERC content with updates); accepted ~Oct 31, 2025; presented Feb 26--28, 2026. |
| **Nov 2025 -- Mar 2026** | Kevin's work shifts to **attribute hierarchies** (e.g., testing whether skills form a strict order P > L > D > AD > I) for the IJRUME journal manuscript and a separate "Skill Hierarchies" paper -- both with their own GitHub repos, distinct from the "Calc Repo." This later work is why `unused_files/Qmatrix_intermediate_coding_stage.xlsx` (39 fine attributes, `CCI_partial`/`CCA_partial`, `PCA_Carlson` variants) looks more evolved than what the conference papers report: **it reflects this later phase, not the 5-LO version the papers actually used.** |

The upshot: there were at least three distinct Q-matrix eras -- (1) an initial
rough coding used for the RUME 2025 submission, (2) the refined, Qval-updated
5-LO coding used for the PERC 2025 paper (**this is the version reconstructed
below**), and (3) ongoing, unfinished attribute-hierarchy work for the IJRUME
and Skill Hierarchies papers.

# The 27-LO taxonomy behind the 5 broad skills

Before narrowing to P/L/D/AD/I, Kevin built out this fuller taxonomy (from
`CCD-second coding.xlsx`, tab `Revised LOs`). Every fine-grained
code below rolls up into exactly one of the five broad skills used in the
final Q-matrix.

```{r lo-taxonomy}
lo_taxonomy <- tribble(
  ~Skill, ~Code, ~Name, ~Description,
  "P", "P1", "Foundations of Functions", "Function notation, evaluation, domain/range, graph, zeros/roots, composition, symmetry, table of values",
  "P", "P2", "Basic Classes of Functions", "Slope/meaning, degree of polynomial, odd/even, rational, power/root, algebraic/transcendental, piecewise, transformations",
  "P", "P3", "Trigonometric Functions", "Degrees/radians, triangle/unit circle, basic trig identities, graphs and periods, transformations",
  "P", "P4", "Inverse Functions", "Conditions/test for invertibility, horizontal line test, find inverse, graph inverse, inverse trig",
  "P", "P5", "Exponential and Logarithmic Functions", "Form of x^b vs b^x, significance of 'e', form of logs, exp vs logs, calculate with logs, hyperbolic functions",
  "P", "P6", "Algebraic Basics", "Inequalities, proportions",
  "P", "P7", "Graphical Fluency", NA,
  "L", "L2", "The Limit of a Function", "Correct notation, limits from table/graph, exist or not, one/two-sided, infinite limit, vertical asymptote",
  "L", "L3", "The Limit Laws", "Basic limit laws, evaluate by factoring, polynomials, rational functions, conjugates, squeeze theorem",
  "L", "L4", "Continuity", "Three conditions for continuity at a point, three kinds of discontinuity, continuity on an interval, limits of compositions, IVT",
  "D", "D1", "Defining the Derivative", "Meaning of tangent line, slope of tangent, derivative as difference quotient, y' at a point, velocity, avg vs instant velocity, y' from a table",
  "D", "D2", "The Derivative as a Function", "Definition of derivative function, graph y' from y, derivatives and continuity, existence, higher-order derivatives",
  "D", "D3", "Differentiation Rules", "Constant, power, product, quotient rules; polynomials, rational functions",
  "D", "D5", "Derivatives of Trig Functions", "Derivatives and higher-order derivatives of sine and cosine (and inverse trig functions)",
  "D", "D6", "The Chain Rule", "State chain rule for f(g(x)); apply with power/product/quotient rule; apply for f(g(h(x))); describe proof",
  "D", "D9", "Derivatives of Exp & Log functions", "Derivative of exponential/log functions; logarithmic differentiation",
  "AD", "AD1", "Related Rates", "Express changing quantities as derivatives, find relationships between derivatives, use chain rule to relate rates",
  "AD", "AD2", "Linear Approximation and Differentials", "Linearization at a point, graphical use of differentials, relative/percentage error",
  "AD", "AD3", "Maxima and Minima", "Absolute/local extrema, finding critical points, open vs. closed interval",
  "AD", "AD5", "Derivatives and shape of a graph", "First/second derivative tests, concavity, inflection points, relationship between y/y'/y''",
  "AD", "AD6", "Limits at Infinity and Asymptotes", "Limits at infinity, horizontal/oblique asymptotes, end behavior, graphing with y'/y''",
  "AD", "AD8", "L'Hopital's Rule", "When to apply, indeterminate forms, relative growth rates of functions",
  "AD", "AD10", "Antiderivatives", "General antiderivative, notation for indefinite integral, power rule, simple IVPs",
  "I", "I1", "Approximating Areas", "Sigma notation, rectangular-area approximation, Riemann sums",
  "I", "I2", "The Definite Integral", "Definition, integrand/limits/variable, integrability, net area, average value of a function",
  "I", "I3", "The Fund. Theorem of Calculus", "MVT for integrals, FTC parts 1 & 2, differentiating integrals, relationship between differentiation and integration",
  "I", "I5", "Substitution", "Substitution for indefinite and definite integrals",
  "I", "I6", "Integrals w/ exponential & logarithms", "Integration of exponential and logarithmic functions"
)
kable(lo_taxonomy)
```

Two codes appear in earlier/other coding tabs but are **not** part of this
27-LO taxonomy or the final Q-matrix: **P6** (Algebraic Basics) and **E1**
("Interpretation") were proposed additions that didn't survive into the final
27-LO/5-skill scheme reconstructed here (P6 appears in the taxonomy above but
with zero items coded to it in the final Q-matrix -- see the item counts
below).

## Item-count expansion roadmap (as of the CCD-second coding.xlsx timeline tabs)

Separately from the Q-matrix itself, the project has a roadmap for growing
item coverage per LO toward a "fixed-length" CCD test, targeting 10 items per
LO by Fall 2027 (`timeline - 10 items` tab; a lower-target `timeline - 6
items` variant also exists). This contextualizes where thin LOs (e.g., P6,
P7, D5/D6/D9, most Integration codes) are headed, even though they currently
have few or zero coded items:

```{r timeline}
timeline <- tribble(
  ~Skill, ~Code, ~Name, ~`Active now`, ~`Test Fall 25`, ~`Test Sp 26`, ~`Test Fall 26`, ~`Test Sp 27`, ~`Total by Fall 27`,
  "P", "P1", "Foundations of Functions", 13, NA, NA, NA, NA, 13,
  "P", "P2", "Basic Classes of Functions", 18, NA, NA, NA, NA, 18,
  "P", "P3", "Trigonometric Functions", 1, 4, 5, NA, NA, 10,
  "P", "P4", "Inverse Functions", 4, NA, NA, 3, 3, 10,
  "P", "P5", "Exponential and Logarithmic Functions", 5, 2, 3, NA, NA, 10,
  "P", "P6", "Algebraic Basics", 0, NA, NA, NA, NA, 0,
  "P", "P7", "Graphical Fluency", 0, NA, NA, NA, NA, 0,
  "L", "L2", "The Limit of a Function", 9, 1, NA, NA, NA, 10,
  "L", "L3", "The Limit Laws", 1, NA, NA, NA, NA, 1,
  "L", "L4", "Continuity", 0, NA, NA, NA, NA, 0,
  "D", "D1", "Defining the Derivative", 4, 3, 3, NA, NA, 10,
  "D", "D2", "The Derivative as a Function", 5, 3, 2, NA, NA, 10,
  "D", "D3", "Differentiation Rules", 0, NA, NA, 5, 5, 10,
  "D", "D5", "Derivatives of Trig Functions", 0, NA, NA, NA, NA, 0,
  "D", "D6", "The Chain Rule", 0, NA, NA, NA, NA, 0,
  "D", "D9", "Derivatives of Exp & Log functions", 0, NA, NA, NA, NA, 0,
  "AD", "AD1", "Related Rates", 1, NA, NA, NA, NA, 1,
  "AD", "AD2", "Linear Approximation and Differentials", 1, NA, NA, NA, NA, 1,
  "AD", "AD3", "Maxima and Minima", 1, NA, NA, 4, 5, 10,
  "AD", "AD5", "Derivatives and shape of a graph", 6, 2, 2, NA, NA, 10,
  "AD", "AD6", "Limits at Infinity and Asymptotes", 3, NA, NA, NA, NA, 3,
  "AD", "AD8", "L'Hopital's Rule", 1, NA, NA, NA, NA, 1,
  "AD", "AD10", "Antiderivatives", 0, NA, NA, NA, NA, 0,
  "I", "I1", "Approximating Areas", 0, NA, NA, 5, 5, 10,
  "I", "I2", "The Definite Integral", 2, 4, 4, NA, NA, 10,
  "I", "I3", "The Fund. Theorem of Calculus", 1, NA, NA, NA, NA, 1,
  "I", "I5", "Substitution", 1, NA, NA, NA, NA, 1,
  "I", "I6", "Integrals w/ exponential & logarithms", 0, NA, NA, NA, NA, 0
)
kable(timeline)
cat("Grand total target across all LOs by Fall 2027: 265 items (10-items-per-LO plan)\n")
```

## PCA's original framework: Carlson (2010)

PCA was not originally designed around the P/L/D/AD/I scheme -- it's Marilyn
Carlson's Precalculus Concept Assessment, built around a different construct
framework, now in `Articles/`:

> Carlson, M., Oehrtman, M., & Engelke, N. (2010). The precalculus concept
> assessment: A tool for assessing students' reasoning abilities and
> understandings. *Cognition and Instruction*, 28(2), 113-145.

`unused_files/Qmatrix_intermediate_coding_stage.xlsx`'s `PCA_Carlson` tab preserves Kevin's coding of all
25 PCA items against Carlson's own 15 sub-construct codes before they were
recoded into P/L/D/AD/I. Those codes are now defined, straight from the
paper's Table 2 ("The PCA Taxonomy"), along with the exact PCA items each
code applies to:

```{r carlson-taxonomy}
carlson_codes <- tribble(
  ~Category, ~Code, ~Name, ~`Items (Carlson et al. 2010, Table 2)`,
  "Reasoning ability", "R1", "Process view of function", "1,2,4,5,6,7,8,9,10,11,12,13,17,20,22,23",
  "Reasoning ability", "R2", "Covariational reasoning", "15,18,19,24,25",
  "Reasoning ability", "R3", "Computational abilities", "1,3,4,10,11,14,16,17,21",
  "Meaning of function concepts", "ME", "Function evaluation", "1,5,6,11,12,16,20",
  "Meaning of function concepts", "MR", "Rate of change", "8,10,11,15,19,22",
  "Meaning of function concepts", "MC", "Function composition", "4,5,12,16,17,20,23",
  "Meaning of function concepts", "MI", "Function inverse", "2,4,9,10,13,14,23",
  "Growth rate of function types", "GL", "Linear", "3,10,22",
  "Growth rate of function types", "GE", "Exponential", "7",
  "Growth rate of function types", "GR", "Rational", "18,25",
  "Growth rate of function types", "GN", "General non-linear", "15,19,24",
  "Function representations", "RG", "Graphical", "2,5,6,8,9,10,15,19,24",
  "Function representations", "RA", "Algebraic", "1,4,7,10,11,14,16,17,18,21,22,23,25",
  "Function representations", "RN", "Numerical", "3,12,13",
  "Function representations", "RC", "Contextual", "3,4,7,8,10,11,15,17,18,20,22"
)
kable(carlson_codes, caption = "Carlson et al. (2010) Table 2: the PCA Taxonomy, item-by-item")
```

**Important:** this is a *different, independent* classification of the same
25 items from the P/L/D/AD/I scheme used in the CDM Q-matrix above -- it's
Carlson's own validated research taxonomy, not something to reconcile
item-for-item against Kevin's later recoding. Both are kept here for
reference, not merged.

### Five known PCA items, verbatim from the source paper

The paper reproduces exact wording and (validated, empirically-derived)
answer choices for five items as worked examples. These are the only PCA
item text known to exist anywhere connected to this project:

**Item 15** (bottle problem; R2 covariation, MR, GN, RG) -- *"Assume that
water is poured into a spherical bottle at a constant rate. Which of the
following graphs best represents the height of water in the bottle as a
function of the amount of water in the bottle?"* Five graphed answer choices
(concave-down/concave-up combinations); correct answer covaries height with
volume (concave down, then concave up).

**Item 17** (circle-ripple problem; R1, R3, MC, RA, RC) -- *"A ball is thrown
into a lake, creating a circular ripple that travels outward at a speed of 5
cm per second. Express the area, A, of the circle in terms of the time, t,
(in seconds) that have passed since the ball hits the lake."* a) A(t)=25*pi*t,
b) A(t)=*pi*t^2^, **c) A(t)=25*pi*t^2^ (correct, ~17%)**, d) A(t)=5*pi*t^2^, e) A(t)=10*pi*t.

**Item 4** (square area/perimeter; R1, R3, MC, MI, RA, RC) -- *"Which of the
following formulas defines the area, A, of a square in terms of its
perimeter, p?"* **a) A=p^2^/16 (correct, ~25%)**, b) A=s^2^, c) A=p^2^/4,
d) A=16x^2^, e) p=4*sqrt*(A).

**Item 11** (average rate of change; R1, R3, ME/MR, RA, RC) -- *"The distance
s (in feet) of a car moving in a straight line is given by s = t^2^ + t,
where t is measured in seconds. Find the average rate of change (velocity)
for the time period from t = 1 to t = 4."* a) 5 ft/sec, **b) 6 ft/sec
(correct, ~15% of 1,501 students)**, c) 9 ft/sec, d) 10 ft/sec, e) 11 ft/sec.

**Item 5** (graphical composition; R1, ME, MC, RG) -- *"Use the graphs of f
and g to evaluate g(f(2))."* Graphs described in-text: f passes through
(-2, 1) and (2, -2); g passes through (4, 3). Final answer choices: a) -2,
**b) 1 (correct, ~43%)**, c) 4, d) 2, e) "not defined" -- the graphs
themselves are figures, not fully recoverable from the extracted text, so
this one is closest to verbatim rather than exact.

None of these five have been cross-referenced against this project's own
PCA item numbering (`pre_1`...`pre_25`) -- that mapping isn't knowable
without either LASSO's item content or Carlson's original instrument
directly, since our PCA item numbers may not match the paper's numbering
1-to-1 (order can differ between the source instrument and how it was
delivered on LASSO).

## CCI's validation history: Epstein (2013)

Also now in `Articles/`:

> Epstein, J. (2013). The Calculus Concept Inventory -- Measurement of the
> Effect of Teaching Methodology in Mathematics. *Notices of the AMS*, 60(8),
> 1018-1026.

This is a methods/validation article, not an item bank -- **it contains no
item text or item-construct mapping** (unlike the Carlson PCA paper above),
but it does explain CCI's origin and validates our item count:

* CCI development began in 2004 (Epstein & Yang, NSF-funded), modeled
  directly on the Force Concept Inventory (FCI) from physics education
  research.
* An expert panel wrote an initial item pool; **"Cognitive Laboratories"**
  (structured think-aloud interviews) confirmed most items hit their
  intended misconceptions, but **the panel removed two items**: one that
  cognitive labs showed wasn't behaving as intended, and one where "we were
  not satisfied that we had found an item to measure the construct of 'the
  derivative as a measure of the sensitivity of the output to change in the
  input' -- that is, essentially the tangent line approximation." **This
  left a final test of twenty-two items** -- confirming our CCI item count
  (22) independently of the LASSO/Q-matrix side.
* Psychometric analysis found roughly two empirical dimensions -- "Functions"
  and "Derivatives" -- plus a smaller third dimension on "limits, ratios, and
  the continuum." This is a coarser, differently-drawn split than the P/L/D/AD/I
  scheme (which separates Limits, Derivatives, and Applications of Derivatives),
  but broadly consistent with it. Reliability (Cronbach's alpha) was 0.7.
* The paper's real focus is validating CCI as a measure of teaching-method
  effect ("normalized gain" `<g>`, following Hake's physics-education
  methodology): interactive-engagement sections consistently showed
  substantially higher gains than traditional lecture sections across many
  institutions -- background that may be useful context for this project's
  own pre/post gain analysis, independent of the Q-matrix question.

# Where the Q-matrix data lives, and why most candidates are wrong

Several files could plausibly be "the Q-matrix," which was the original
source of confusion this document exists to resolve. Only one is:

| File | What it actually is |
|---|---|
| `Qmatrix.xlsx` (repo root) | **The Q-matrix `IJRUME_analysis.R`/`.Rmd` actually read.** The reconstructed final Q-matrix, **generated by this document** -- see below. |
| `CCD-second coding.xlsx` (repo root) | **The source of truth this reconstruction is derived from.** Sheets `CCI Q`, `CCA Q`, `PCA Q` hold the final, Qval-updated fine-grained coding (11/11/9 attributes respectively) that collapses to the 5-skill scheme used in the papers. |
| `unused_files/Qmatrix_intermediate_coding_stage.xlsx` | An **older, intermediate coding stage**, not read by any analysis script: the ~37-code fine-grained taxonomy (P1-P7, L1-L5, D1-D9, AD1-AD10, I1-I7, E1) from Kevin's later (Nov 2025+) attribute-hierarchy work, plus the Carlson-framework tabs. Sheet names are uppercase (`CCI`/`CCA`/`PCA`), which doesn't even match what `IJRUME_analysis.R` expects (`sheet = "cci"` etc., case-sensitive). Kept for historical reference, not deleted, since it's discussed throughout this document's [History: how we got to a 5-skill Q-matrix] section. |

Two other files were in earlier versions of this repo and have since been
removed entirely (not just moved) because they were neither used by any
script nor useful context: a root-level `Qmatrix.csv` (20 items x 8 generic
attributes named "Item01"/"A1" -- the classic Tatsuoka fraction-subtraction
dataset bundled as a tutorial example with CDM packages, unrelated to this
project), and `Copy_Qmatrix.xlsx` (a byte-for-byte duplicate of the
intermediate coding stage above).

# Reconstructing the 5-skill Q-matrix

```{r load-source}
source_path <- "CCD-second coding.xlsx"
stopifnot(file.exists(source_path))

read_fine_qmatrix <- function(sheet) {
  raw <- suppressMessages(read_excel(source_path, sheet = sheet, col_names = FALSE))
  header_row <- as.character(unlist(raw[2, ]))
  item_cols <- which(!is.na(suppressWarnings(as.numeric(header_row))))
  items <- as.numeric(header_row[item_cols])
  attr_rows <- raw[-(1:2), ]
  attr_rows <- attr_rows[!is.na(attr_rows[[1]]), ]
  attrs <- as.character(attr_rows[[1]])
  mat <- as.data.frame(attr_rows[, item_cols])
  mat[] <- lapply(mat, function(x) as.numeric(x) %in% 1)
  colnames(mat) <- paste0("item", items)
  rownames(mat) <- attrs
  mat
}

collapse_to_skills <- function(fine_mat, skills = c("P", "L", "D", "AD", "I")) {
  attr_names <- rownames(fine_mat)
  skill_of <- function(a) {
    m <- regmatches(a, regexpr("^[A-Za-z]+", a))
    m
  }
  attr_skill <- vapply(attr_names, skill_of, character(1))
  out <- sapply(skills, function(s) {
    rows <- fine_mat[attr_skill == s, , drop = FALSE]
    if (nrow(rows) == 0) return(rep(0L, ncol(fine_mat)))
    as.integer(colSums(rows, na.rm = TRUE) > 0)
  })
  out <- as.data.frame(out)
  colnames(out) <- skills
  rownames(out) <- colnames(fine_mat)
  out
}

fine_cca <- read_fine_qmatrix("CCA Q")
fine_cci <- read_fine_qmatrix("CCI Q")
fine_pca <- read_fine_qmatrix("PCA Q")

q_cca <- collapse_to_skills(fine_cca)
q_cci <- collapse_to_skills(fine_cci)
q_pca <- collapse_to_skills(fine_pca)
```

The fine-grained source coding, before collapsing (rows = the specific LOs
actually coded to at least one item; columns = item number):

```{r show-fine, results='asis'}
cat("**CCA** fine attributes used:", paste(rownames(fine_cca), collapse = ", "), "\n\n")
cat("**CCI** fine attributes used:", paste(rownames(fine_cci), collapse = ", "), "\n\n")
cat("**PCA** fine attributes used:", paste(rownames(fine_pca), collapse = ", "), "\n\n")
cat("(PCA has no 'D' fine attribute coded to any item -- consistent with the",
    "manuscripts' note that no PCA items were coded as assessing Derivatives.)\n")
```

Collapsed to the 5 broad skills (item x skill, 1 = this item counts toward
mastery of that skill):

```{r show-collapsed}
kable(tibble(item = rownames(q_cca), q_cca), caption = "CCA (17 items)")
kable(tibble(item = rownames(q_cci), q_cci), caption = "CCI (22 items)")
kable(tibble(item = rownames(q_pca), q_pca[, c("P","L","AD","I")]), caption = "PCA (25 items; no Derivatives-coded items)")
```

# Verification against the published manuscript

Table IV of `Articles/PERC2025_Roberge.pdf` reports the number of items per
skill for each instrument's *final* Q-matrix. If this reconstruction is
correct, collapsing the fine-grained coding above should reproduce those
counts exactly.

```{r verify}
published_table_iv <- tribble(
  ~Instrument, ~P, ~L, ~D, ~AD, ~I,
  "CCA", 7, 4, 3, 2, 2,
  "CCI", 6, 3, 6, 6, 1,
  "PCA", 20, 3, 0, 5, 1
)

reconstructed_counts <- bind_rows(
  c(Instrument = "CCA", colSums(q_cca)),
  c(Instrument = "CCI", colSums(q_cci)),
  c(Instrument = "PCA", colSums(q_pca))
) %>% mutate(across(c(P, L, D, AD, I), as.integer))

comparison <- bind_rows(
  published_table_iv %>% mutate(source = "Published (Table IV)"),
  reconstructed_counts %>% mutate(source = "Reconstructed here")
) %>% arrange(Instrument, source)

kable(comparison, caption = "Items-per-skill: published vs. reconstructed")

match_check <- all(
  reconstructed_counts %>% arrange(Instrument) %>% select(P, L, D, AD, I) ==
  published_table_iv %>% arrange(Instrument) %>% select(P, L, D, AD, I)
)
cat(if (match_check) "MATCH: reconstruction reproduces Table IV exactly.\n"
    else "MISMATCH -- investigate before trusting Qmatrix.xlsx.\n")

n_items_check <- tribble(
  ~Instrument, ~n_items, ~paper_says,
  "CCA", nrow(q_cca), 17,
  "CCI", nrow(q_cci), 22,
  "PCA", nrow(q_pca), 25
)
kable(n_items_check, caption = "Item count check")

double_coded <- tibble(
  Instrument = c("CCA", "CCI", "PCA"),
  n_double_coded = c(sum(rowSums(q_cca) > 1), sum(rowSums(q_cci) > 1), sum(rowSums(q_pca) > 1))
)
kable(double_coded, caption = "Items coded to more than one skill (paper says 5 total across all three instruments)")
cat("Total double-coded:", sum(double_coded$n_double_coded), "(paper says 5)\n")
```

This reconstruction also reproduces the item numbering gaps the R scripts
depend on: CCI skips items 1-2 (starts at item 3), CCA skips item 12 -- both
match the grading-key item lists already hard-coded in `IJRUME_analysis.R`.

# Generating Qmatrix.xlsx

This chunk writes the workbook `IJRUME_analysis.R`/`.Rmd` actually read, in
the exact shape they expect: sheets `pca`/`cca`/`cci` (lowercase), an `LO`
item-label column, a `D` placeholder metadata column for `pca` only (dropped
by the analysis code, present only because the original script's
`select(-LO, -D)` expects it), and the skill columns. It's off by default
(`params$write_qmatrix`) so rendering this document for reading doesn't
silently overwrite the repo's `Qmatrix.xlsx` -- set `write_qmatrix: true` to
regenerate it.

```{r write-qmatrix, eval=write_qmatrix}
readme <- tibble(
  field = c("Purpose", "Source", "Verification", "Coding origin", "Skills", "Sheet shapes", "Caveat"),
  value = c(
    "Reconstructed final Q-matrix for the CCA/CCI/PCA DINA/GDINA models in IJRUME_analysis.R, generated by Qmatrix_documentation.Rmd.",
    "Derived from CCD-second coding.xlsx, sheets 'CCA Q', 'CCI Q', 'PCA Q' (fine-grained attributes), collapsed to the 5 broad categories P/L/D/AD/I.",
    "Item counts per skill (P/L/D/AD/I) reproduce Table IV of Articles/PERC2025_Roberge.pdf exactly -- see Qmatrix_documentation.Rmd for the check.",
    "Per Roberge_agendas: coded by Kevin Roberge with review from Jayson Nissen; the '5 LO, Qval-updated' version (1 change to CCA, 2 changes to CCI) used for the PERC2025/RUME manuscripts.",
    "P=Prerequisite skills, L=Limits, D=Derivatives, AD=Applications of Derivatives, I=Integration",
    "pca: 25 items x [D placeholder, P, L, AD, I]; cca: 17 items x [P, L, D, AD, I]; cci: 22 items x [P, L, D, AD, I].",
    "Does NOT reflect Kevin Roberge's later attribute-hierarchy work (Nov 2025-Mar 2026) in unused_files/Qmatrix_intermediate_coding_stage.xlsx -- confirm with Kevin/Jayson before using for new analysis."
  )
)

make_sheet <- function(q, extra_placeholder = NULL) {
  items <- as.integer(sub("^item", "", rownames(q)))
  out <- tibble(LO = paste0("item", items))
  if (!is.null(extra_placeholder)) out[[extra_placeholder]] <- NA
  bind_cols(out, q[order(items), , drop = FALSE])
}

write_xlsx(
  list(
    README = readme,
    pca = make_sheet(q_pca[, c("P","L","AD","I")], extra_placeholder = "D"),
    cca = make_sheet(q_cca),
    cci = make_sheet(q_cci)
  ),
  path = "Qmatrix.xlsx"
)
cat("Wrote Qmatrix.xlsx\n")
```

# Item key: numbers, skills, and (missing) content

```{r item-key}
make_key <- function(q, instrument) {
  tibble(instrument = instrument, item = rownames(q)) %>%
    bind_cols(select(q, any_of(c("P","L","D","AD","I")))) %>%
    mutate(
      skills = pmap_chr(select(., any_of(c("P","L","D","AD","I"))), function(...) {
        vals <- c(...)
        paste(names(vals)[vals == 1], collapse = "+")
      }),
      item_text = NA_character_
    )
}
item_key <- bind_rows(
  make_key(q_cca, "CCA"),
  make_key(q_cci, "CCI"),
  make_key(q_pca, "PCA")
) %>% select(instrument, item, skills, item_text)

kable(item_key, caption = "Full item-skill key. item_text is blank everywhere -- see below.")
```

## What's missing: item text

**Almost no item wording exists in this repository, for either CCA or CCI or
the bulk of PCA.** The exceptions are five PCA items (4, 5, 11, 15, 17) with
verbatim wording in Carlson et al. (2010), reproduced above -- and even those
aren't confirmed to line up with this project's own item numbering (see
caveat above). Confirmed by:

* LASSO's API (`/api/assessments/{year}/regular/export`, the only endpoint
  this project's code uses) returns item numbers, item-bank codes, response
  letters, timing, and scores -- never question text.
* No codebook file sits alongside `pca_8_23_df.csv` or any LASSO download.
* `CCD-second coding.xlsx`'s coding tabs (`Coding-KR`, `Coding-JN`,
  `2ndCoding Qs`, `CCI Q`/`CCA Q`/`PCA Q`) are binary item-by-LO flag
  matrices -- item numbers only, never item text.
* Epstein (2013), CCI's validation paper, has no item-level content either
  (it's a methods/validity article, not an item bank).

Filling this in is still unfinished work -- see the "Known open issues"
section of the repo's top-level `README.md`. Test items are confidential
and are intentionally not reproduced, described, or extracted into this
(public) repository beyond the already-published excerpts from Carlson
(2010) above.

**Next steps:**

1. Confirm PCA's data provenance (see next section) -- whoever originally
   extracted `pca_8_23_df.csv` may also have item-level documentation.
2. ~~Track down Carlson, Oehrtman & Engelke (2010)~~ -- done; see above.

## PCA's data provenance is still unresolved

`pca_8_23_df.csv` is not fetched by this project's LASSO download pipeline,
and `download_lasso_data.Rmd` itself flags this as an open question. But its
metadata (exactly 8 institutions, years 2015-2018) matches
`Articles/PERC2025_Roberge.pdf`'s stated PCA provenance exactly, so it likely
*did* originate from LASSO, extracted separately (and earlier) by Kevin
Roberge or a Purdue collaborator (Yuxiao Zhang, per the agendas), possibly
under a LASSO assessment code that's since changed, or via an export type
(`likert`/`adaptive`) this project has never tried. This needs confirmation
from Jayson, not further guessing from the data alone.

# Open questions (full list)

1. Confirm this reconstruction (matching Table IV exactly, above) is the Q-matrix version to build on for IJRUME -- not the later attribute-hierarchy work in `unused_files/Qmatrix_intermediate_coding_stage.xlsx`.
2. What does CCA item 12 actually measure? The bogus `== "F"` eligibility filter built on it is fixed (2026-07-23, see `datascript2.Rmd`), but the underlying question -- what item 12 itself is checking -- is still open. (Unrelated to this Q-matrix.)
3. Should `C1CI`/`C2CI` (Calculus 1/2 Concept Inventory codes on LASSO) be incorporated for more CCI coverage, or are they a different instrument version?
4. Where can we get the rest of PCA's/CCI's/CCA's item content, and can Jayson confirm LASSO's original PCA export still exists?
5. Is LASSO's pre-2025 missing-scores bug fixed yet? (Confirmed still broken as of 2026-07-22.)
6. Should the CDM (DINA/GDINA) fitting enforce pre/post measurement invariance, like the separate 3PL/IRT section already does?
7. ~~Get a copy of Carlson, Oehrtman & Engelke (2010)~~ -- done (2026-07-22); its Table 2 is now reproduced above. Epstein (2013), CCI's validation paper, is also now in `Articles/` -- also incorporated above, though it has no item-level content.
8. Do the five known PCA item numbers from Carlson (2010) (items 4, 5, 11, 15, 17) actually match this project's `pre_4`/`pre_5`/`pre_11`/`pre_15`/`pre_17` numbering on LASSO? Not yet confirmed.
